---
title: "03_WGCNA_network_analysis"
output: html_document
date: "2025-03-14"
---
# Setup
Load packages
```{r}
library(tidyverse)
library(DESeq2)
library(data.table)
library(WGCNA) #BiocManager::install("impute")
allowWGCNAThreads() #allow multi-threading


dir <- "/workdir/ssv42/RNAseq/"
dir2 <- "/workdir/ssv42/RNAseq_yr2/"
set.seed(14850)
```

Get partitioned gene expression estimates
Get LFC coefficient matrix from DESeq2 & extract genotypic, non-genotypic, and overall (SV-adjusted) estimates
```{r}
deseq_wsv_analysis <- readRDS(file=paste0(dir2, "output/DEseq_analysis_w7SVfactors.RDS"))
LFCest <- readRDS(file=paste0(dir2, "output/coefficients_from7SVmodel_log2FC_estimates.RDS")) %>% as.data.frame()

#extract genotype coefficients ("expression BLUEs"), add intercept to every column
genoLFCs <- LFCest[,c(1, grep("genotype", colnames(LFCest)))] %>%
  mutate(across(-1, ~ .x + Intercept)) %>%
  dplyr::rename("genotype_A03" = "Intercept") %>%
  dplyr::rename_with(~ gsub("_vs_A03", "", .)  )

saveRDS(genoLFCs, file= paste0(dir2, "output/genotypicLFCestimates.RDS") )



#extract non-genotypic coefficients (sum of every other coefficient besides genotypes and SVs, plus residuals)
log2residuals <- readRDS(file=paste0(dir2, "output/residuals_from7SVmodel_log2scale_estimates.RDS"))
metadata_phenos_scaled_sv <- readRDS(file=paste0(dir2, "output/metadata_phenos_scaled_wsvs.RDS"))
model_matrix <- model.matrix(deseq_wsv_analysis@design, metadata_phenos_scaled_sv)
rownames(model_matrix) <- metadata_phenos_scaled_sv$fullID

#confirm matching sample order in residuals and model matrix dataframes:
all(colnames(log2residuals) == rownames(model_matrix))

#calculate non-genotypic estimates
nongeno_model_matrix <- as.data.frame(model_matrix) %>%  dplyr::select(!contains("genotype")) %>%
  dplyr::select(-c(V1:V7)) 
nongeno_coefs <- LFCest  %>%  dplyr::select(!contains("genotype")) %>% dplyr::select(-c(V1:V7))
nongeno_estimates <- (as.matrix(nongeno_coefs) %*% t(nongeno_model_matrix)) + log2residuals

saveRDS(nongeno_estimates, file= paste0(dir2, "output/nongenotypicLFCestimates.RDS") )



# calculate overall estimates (adjusted for SVs)
overall_model_matrix <- as.data.frame(model_matrix) %>%  dplyr::select(-c(V1:V7)) 
overall_coefs <- LFCest  %>%  dplyr::select(-c(V1:V7))
overall_estimates <- (as.matrix(overall_coefs) %*% t(overall_model_matrix)) + log2residuals

saveRDS(overall_estimates, file= paste0(dir2, "output/overallLFCestimates.RDS") )

```



# Get partitioned metabolite estimates
```{r, eval=F}
metadata_phenos_scaled <- readRDS(file=paste0(dir2, "output/scaled_pheno_data.RDS"))

#Carotenoid BLUEs
carot_data <- metadata_phenos_scaled %>% dplyr::select(fullID:harvest_date, carot_extraction_date, 
                                    Phytoene:Zeta.Carotene, Violaxanthin.peak.1:Lutein,
                                    total_carotenoids:total_violaxanthin,
                                    percent_bcarotene:ratio_viola_bc,
                                    i_o_diff_TC) %>%
  mutate(root = as.factor(root), block = as.factor(block)) 

#pull out carotenoid compound/trait names
carot_traits <- colnames(carot_data)[13:ncol(carot_data)]

#initiate BLUE storage
carot_trait_coefs <- matrix(nrow=62, ncol = length(carot_traits))

#loop through each trait and fit linear model
for(i in seq_along(carot_traits)){
  
  traiti <- carot_traits[i]
  print(traiti)
  
  traiti_mod <- lm( formula( paste0(traiti, "~ 1 + genotype + block + layer + root + year:timept + carot_extraction_date") ),  
                      data = carot_data)
  
  traiti_predictions <- predict(traiti_mod, type="terms", terms=c("genotype"))
  
  #extract coefficients and save to matrix
  traiti_coefs <- traiti_mod$coefficients
  carot_trait_coefs[,i] <- traiti_coefs
    # get BLUEs:
    #traiti_BLUEs <- traiti_mod$coefficients[c(1, grep("genotype", names( traiti_mod$coefficients)))] 
    #carot_BLUEs[i,] <- traiti_BLUEs
}

carot_BLUEs <- as.data.frame(carot_BLUEs) 
colnames(carot_BLUEs) <- names(traiti_BLUEs)
rownames(carot_BLUEs) <- carot_traits

#Calculate dry matter BLUEs:
dm_mod <- lm( dry_matter ~ 1 + genotype + year:timept + layer + block + dry_date,  
                      data = metadata_phenos_scaled)
dm_io_mod <- lm( i_o_diff_DM ~ 1 + genotype + year:timept + layer + block + dry_date,  
                      data = metadata_phenos_scaled)

dm_BLUEs <- rbind(dm_mod$coefficients[c(1, grep("genotype", names( dm_mod$coefficients )))],
                  dm_io_mod$coefficients[c(1, grep("genotype", names( dm_io_mod$coefficients )))])


#Merge with carotenoid BLUEs:
allpheno_BLUEs <- rbind(carot_BLUEs, dm_BLUEs)
rownames(allpheno_BLUEs)[c(nrow(allpheno_BLUEs)-1, nrow(allpheno_BLUEs))] <- c("dry_matter", "i_o_diff_DM")

#add intercept to every column
allpheno_BLUEs_int <- allpheno_BLUEs %>% mutate(across(-1, ~ .x + allpheno_BLUEs$`(Intercept)`))
colnames(allpheno_BLUEs_int)[1] <- "genotypeA03"

#save phenotype BLUEs
saveRDS(allpheno_BLUEs_int, file=paste0(dir2, "output/PhenotypeBLUEs.RDS"))
```



# Make overall network
Read in overall estimates (SV-adjusted, log2-normalized count data):
```{r}
#read in variance-transformed normalized count data for overall network (year and 5 SVs regressed out)
overall_estimates <- readRDS(file=paste0(dir2, "output/overallLFCestimates.RDS"))

#spot check to confirm approximate normal distribution of log2 count data: looks good
#hist(overall_estimates[15,])


#transpose so genotypes in rows and genes in columns
overallcounts_t <- t(overall_estimates)

#check everything is numeric
all(apply(overallcounts_t, 2, is.numeric))

#check variance across samples
gene_vars <- apply(overallcounts_t, 2, var)
hist(gene_vars, breaks=100, main="distribution of gene variance")
#gene_vars[1:7] #some have NaNs -- why?
# I have -Inf in my counts data
# confirm all -Inf are removed after adding a small positive value to the zero counts in residuals calculation
sum(is.infinite(gene_vars)) == 0

#manually check gene with very high variance to see if I need to remove
hist(overallcounts_t[,which.max(gene_vars)], breaks=50, xlab="SV-corrected log2 counts", 
     main="expression of gene with highest \nvariance across samples")

#alternative approach: replace infinite values with NA:
#overallcounts_t[is.infinite(overallcounts_t)] <- NA
#any(is.infinite(overallcounts_t))  # should be FALSE now


#check for outliers with WGCNA function:
gsg <-goodSamplesGenes(overallcounts_t, verbose=3)
gsg$allOK
#looks good
```

Pick power for network correlations
The “power” parameter defines the exponent used when transforming the correlation values, which affects number of modules identified
```{r, eval=FALSE}
# Choose a set of soft-thresholding powers
powers = c(1:20) #seq(from = 1, to = 40, by = 5)

allowWGCNAThreads() 
# Call the network topology analysis function
sft = pickSoftThreshold(
  overallcounts_t,   # <= Input data
  powerVector = powers,
  verbose = 5,
  networkType = "unsigned",
  corFnc="bicor")

saveRDS(sft, file=paste0(dir2, "output/picksoftthres_overall.RDS"))
```

Plot power threshold graphs
```{r}
sft <- readRDS(paste0(dir2, "output/picksoftthres_overall.RDS"))

par(mfrow = c(1,2)); cex1 = 0.9

#png(paste0(dir2, "figures/overallnetwork_softthresholdpower_R2.png"))
  plot(sft$fitIndices[, 1],
       -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       xlab = "Soft Threshold (power)",
       ylab = "Scale Free Topology Model Fit, signed R^2",
       main = paste("Scale independence")
  )
  text(sft$fitIndices[, 1],
       -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       labels = powers, cex = cex1, col = "red"
  )
  abline(h = 0.8, col = "red")
  abline(h = 0.9, col = "red")
#dev.off()


#png(paste0(dir2, "figures/overallnetwork_softthresholdpower_connectivity.png"))
  plot(sft$fitIndices[, 1],
       sft$fitIndices[, 5],
       xlab = "Soft Threshold (power)",
       ylab = "Mean Connectivity",
       type = "n",
       main = paste("Mean connectivity")
  )
  text(sft$fitIndices[, 1],
       sft$fitIndices[, 5],
       labels = powers,
       cex = cex1, col = "red")
#dev.off()

```

Construct overall network with WGCNA
Topological Overlap Matrix (TOM) is a function of the adjacency matrix (correlation-based) and connectivity (sum of adjacencies), adding information about the context of the genes within the cluster, to be more robust to spurious correlations.
High TOM = genes are directly connected and/or they share many neighbors.
TOM-based dissimilarity (1 - TOM) is then used for clustering genes into modules.
```{r}
picked_power <- 6
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
overall_netwk <- blockwiseModules(overallcounts_t,                
                          
                          # == Adjacency Function ==
                          power = picked_power,       
                          networkType = "unsigned",   #includes both positive and negative correlations
                          corType = "bicor",  #more robust to outliers than pearson correlation
                          nThreads = 38,

                          # == Tree and Block Options ==
                          randomSeed = 14850,
                          deepSplit = 3, #value between 0 and 4, sensitivity to module splitting
                          pamRespectsDendro = F,
                          minModuleSize = 30,
                          maxBlockSize = ncol(overallcounts_t), #allow for running in 1 block since we have enough computational power

                          # == Module Adjustments ==
                          reassignThreshold = 0.05,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)            
                          saveTOMs = T,
                          saveTOMFileBase = "/workdir/ssv42/RNAseq_yr2/output/overall_ntwk_pwr6_",

                          # == Output Options
                          numericLabels = T,
                          verbose = 2)

cor <- temp_cor # Return original cor function 

saveRDS(overall_netwk, file=paste0(dir2, "output/overall_netwk.RDS"))
```

Look at the overall network modules
```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(overall_netwk$colors)

blockselection <- 1 

# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  overall_netwk$dendrograms[[blockselection]],
  mergedColors[overall_netwk$blockGenes[[blockselection]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )

ggsave(file=paste0(dir2, "figures/overall_network_dendrogram_block", blockselection, ".png"))


overall_module_df <- data.frame(
  gene_id = names(overall_netwk$colors),
  colors = labels2colors(overall_netwk$colors)
)

#Get TOM
load(paste0(overall_netwk$TOMFiles))
overall_TOM <- as.matrix(TOM)


#export gene overall_module_df
write_delim(overall_module_df,
            file = paste0(dir2, "output/overall_network_gene_modules.txt"),
            delim = "\t")
```

Test trait/module associations for overall network
Correlate module eigengenes with phenotypes (centered and scaled)
```{r}
# Get Module Eigengenes per cluster
MEs0 <- moduleEigengenes(overallcounts_t, mergedColors)$eigengenes

# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0);
#colnames(MEs0) <- gsub("ME", "", colnames(MEs0))
module_order = names(MEs0) %>% gsub("ME", "", .)

# Read in phenotypes
metadata_phenos_scaled <- readRDS(file=paste0(dir2, "output/scaled_pheno_data.RDS"))
#confirm sample order is the same: 
all(metadata_phenos_scaled$fullID == rownames(MEs0))
#combine with module eigengenes 
MEs0_wphenos <- cbind(MEs0,  
    (metadata_phenos_scaled %>% dplyr::select(Phytoene:Zeta.Carotene, Violaxanthin.peak.1:Lutein,
                        total_carotenoids:dry_matter, i_o_diff_TC))  ) 

# calculate correlations
MEs0_wphenos_cormat <- cor(MEs0_wphenos, use="pairwise.complete.obs")
MEs0_wphenos_cormat <- as.data.frame(MEs0_wphenos_cormat) 
MEs0_wphenos_cormat$module <- rownames(MEs0_wphenos_cormat)
MEs_wphenos_cormat <- MEs0_wphenos_cormat %>% pivot_longer(-module, names_to="phenotype") %>%
  filter(!str_detect(phenotype, "ME")) %>% 
  filter(str_detect(module, "ME")) %>%
  mutate(module = gsub("ME","", module)) %>%
  mutate(phenotype = factor(phenotype, levels=c("Phytoene", "Phytofluene", "Zeta.Carotene", "Beta.Carotene", "Lutein", 
                                     "Violaxanthin.peak.1", "Violaxanthin.peak.2", "total_carotenoids",
                                    "totalpre", "total_violaxanthin", "ratio_bc_v", "i_o_diff_TC", "dry_matter")) )


ggplot(MEs_wphenos_cormat, aes(y=phenotype, x=module, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Module-trait Relationships from Overall Network", y = "Modules", fill="corr")

ggsave(filename=paste0(dir2, "figures/overallntwk_module_trait_correlations.png"),
       height=6, width=18, units="in")

```

Identify carotenoid and carbohydrate genes 
```{r}
#Get GOterms for a priori candidate genes involved in carotenoid and carbohydrate metabolism
carotPWY <- fread("./Mesc671_v8_assembly/annotation/Carotenoid_metabolism_PWYannotations.csv") %>%
  distinct(.)
carotGO <- fread("./Mesc671_v8_assembly/annotation/Carotenoid_metabolism_GOannotations.csv") %>%
  distinct(.)
carbPWY <- fread("./Mesc671_v8_assembly/annotation/Carbohydrate_metabolism_PWYannotations.csv") %>%
  distinct(.)
carbGO <- fread("./Mesc671_v8_assembly/annotation/Carbohydrate_metabolism_GOannotations.csv") %>%
  distinct(.)


#read in genome annotations
v8_gene_annotations <- fread("./Mesc671_v8_assembly/annotation/Mesculenta_671_v8.1.annotation_info.txt")
pthways <- fread("./Mesc671_v8_assembly/annotation/Mesculenta_v8.1_PWY_desc_annotations.txt")

GO_wide <-  str_split(v8_gene_annotations$GO, " ", simplify=T)
colnames(GO_wide) <- paste0("GO_", 1:ncol(GO_wide))

fullgene_annotations <- cbind(v8_gene_annotations, GO_wide) %>%
  pivot_longer(cols= GO_1:GO_13, names_to="GOtermnumber", values_to="GOterm") %>%
  left_join(car_GOterms, by = "GOterm") %>%
  distinct(locusName,GOterm, .keep_all=TRUE) %>%
  left_join(., pthways, by=c("locusName" = "Gene Name")) %>%
  mutate(class = case_when( `Pathway ID` %in% carotPWY$Pathways  ~ "carot",
                           `Pathway ID` %in% carbPWY$Pathway ~ "carb",
                           GOterm %in% carotGO$Goterm ~ "carot",
                           GOterm %in% carbGO$GOTerm ~ "carb",
                           T ~ NA)) %>% 
  group_by(locusName) %>% summarise(
    Pathways = str_c(unique(`Pathway ID`), collapse = "; "),
    class = unique(str_c(na.omit(class)))[1],
    across( c(Description, GO, KOG, KO, Pfam, Panther, `Best-hit-arabi-defline`), ~ first(unique(.x))) )
    
saveRDS(fullgene_annotations, file=paste0(dir2, "output/fullgene_annotations_wcarot-carb_class.RDS"))
write.csv((fullgene_annotations %>% mutate(Description = gsub(",",";",Description))), 
          file=paste0(paste0(dir2, "output/fullgene_annotations_wcarot-carb_class.csv")))


gene_car_membership <- fullgene_annotations %>%
  distinct(locusName, Description, class ) %>%
  filter(class != "NA")

saveRDS(gene_car_membership, file=paste0(dir2, "output/carotenoid_carbohydrate_gene_list.RDS"))


#check how many genes in each class
table(gene_car_membership$class)


# manually check pathway coverage
#View(overall_modules_annot %>% filter(class == "carot"))
#View(overall_modules_annot %>% filter(class == "carb"))
```

Merge module results with gene annotations to test how often carot/carb genes cluster together:
```{r}
#merge annotations with overall network module dataframe
overall_modules_annot <- overall_module_df %>% 
  mutate(locusName = gsub(".v8.1", "", gene_id)) %>%
  left_join(., fullgene_annotations)

#check how many genes in each class
table(overall_modules_annot$class)

#check which modules carotenoid genes are in
table(overall_modules_annot %>% filter(class=="carot") %>% dplyr::select(colors) )

#check carotenoid superpathway genes specifically
table(overall_modules_annot %>% filter(grepl("CAROTENOID-PWY", Pathways)) %>% dplyr::select(colors))
overall_modules_annot %>% filter(grepl("CAROTENOID-PWY", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

#check which modules carbohydrate genes are in
table(overall_modules_annot %>% filter(class=="carb") %>% dplyr::select(colors) )

#check starch synthesis pathway genes specifically
table(overall_modules_annot %>% filter(grepl("PWY-622", Pathways)) %>% dplyr::select(colors) )
overall_modules_annot %>% filter(grepl("PWY-622", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

#check starch degradation pathway genes specifically
table(overall_modules_annot %>% filter(grepl("PWY-6724", Pathways)) %>% dplyr::select(colors) )
overall_modules_annot %>% filter(grepl("PWY-6724", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

```

Select modules of interest from overall network and export for Cytoscape
```{r}
#check module sizes
table(overall_module_df$colors)

#identify modules most associated with the phenotypes
top_associations <- MEs_wphenos_cormat %>% group_by(phenotype) %>%
  slice_max(abs(value), n=3) 

top_modules <- unique((top_associations %>% filter(abs(value) > 0.5))$module)
print(top_modules)

overall_modules_of_interest <- "turquoise" #top_modules

# Pull out list of genes in that module
submod_overall <- overall_module_df %>%
  subset(colors %in% overall_modules_of_interest)

row.names(overall_module_df) = overall_module_df$gene_id

# Subset TOM matrix for modules of interest
colnames(overall_TOM) <- overall_module_df$gene_id
rownames(overall_TOM) <- overall_module_df$gene_id

subset_overallTOM <- overall_TOM[rownames(overall_TOM) %in% submod_overall$gene_id,
                                colnames(overall_TOM) %in% submod_overall$gene_id]

# check dimensions
dim(subset_overallTOM)
format(object.size(subset_overallTOM), units="GB")

#export subset for Cytoscape
Cyexport <- exportNetworkToCytoscape(subset_overallTOM,
                         edgeFile = paste0(dir2, "output/overallntwk_modulesofinterest_CytoscapeInput-edges_turquoise.txt"),
                         nodeFile = paste0(dir2, "output/overallntwk_modulesofinterest_CytoscapeInput-nodes_turquoise.txt"),
                         weighted = TRUE, threshold = 0.1,
                         nodeNames = (colnames(subset_overallTOM) %>% gsub(".v8.1","",.)) ) #remove version tag from gene name

#export gene annotations for Cytoscape
subset_overallTOM_annot <- overall_modules_annot %>% 
  filter(gene_id %in% unique(c(rownames(subset_overallTOM), colnames(subset_overallTOM)))) %>%
  dplyr::select(gene_id, locusName, Description, colors, class) %>%
  mutate(Description = gsub(",",";",Description)) #remove commas to not confuse cytoscape reading csv

write.csv(subset_overallTOM_annot, file= paste0(dir2, "output/overallntwk_modulesofinterest_CytoscapeInput-nodes_annotations_turquoise.csv"), row.names = F, quote=F)
```

Create subnetwork of only carotenoid and carbohydrate genes
```{r}
car_gene_subset <- overall_modules_annot %>% filter(!is.na(class)) %>%
  mutate(Description = gsub(",",";",Description))

# Subset TOM matrix for genes of interest
overallTOM_car_genes <- overall_TOM[rownames(overall_TOM) %in% car_gene_subset$gene_id,
                                colnames(overall_TOM) %in% car_gene_subset$gene_id]

#export subset for Cytoscape
Cyexport <- exportNetworkToCytoscape(overallTOM_car_genes,
                         edgeFile = paste0(dir2, "output/overallntwk_modulesofinterest_CytoscapeInput-edges_cargenes.txt"),
                         nodeFile = paste0(dir2, "output/overallntwk_modulesofinterest_CytoscapeInput-nodes_cargenes.txt"),
                         weighted = TRUE, threshold = 0.01,
                         nodeNames = (colnames(overallTOM_car_genes) %>% gsub(".v8.1","",.)) ) #remove version tag from gene name

write.csv(car_gene_subset, file=paste0(dir2, "output/overallntwk_modulesofinterest_CytoscapeInput_cargene_annotations.csv"))
```

Calculate correlation matrix separately on genes of interest subset to preserve sign/direction information
```{r}
overall_subset_bicor <- bicor(
  overallcounts_t[,colnames(overallcounts_t) %in%  car_gene_subset$gene_id] ) %>% 
  reshape2::melt() 

colnames(overall_subset_bicor) <- c("sourceNode", "targetNode", "bicor")

overall_subset_bicor_Cyto <- overall_subset_bicor %>%
  mutate(across(sourceNode:targetNode, ~ gsub(".v8.1", "", .))) %>%
  mutate(shared_name = paste0(sourceNode, " (interacts with) ", targetNode)) %>%
  mutate(cor_direction = case_when(bicor > 0 ~ "+", bicor <= 0 ~ "-"))

write.csv(overall_subset_bicor_Cyto, 
          file=paste0(dir2, "output/overallnet_cargene-subset_bicorrelations.csv"), 
          row.names = F)
```

What is the relationship between TOM value and bicor?
```{r}
test <- overallTOM_car_genes %>% reshape2::melt() %>%
  mutate(sourceNode = gsub(".v8.1","",Var1), targetNode = gsub(".v8.1","",Var2)) %>%
  left_join(overall_subset_bicor_Cyto)

plot(test$bicor, test$value, xlab="Biweight Midcorrelation", ylab="TOM")

```

Identify hub genes:
```{r}
#hubGeneSignificance(datKME, GS)

#chooseTopHubInEachModule
```

#### end of overall network ####







# Construct non-genetic network
Read in non-genotypic expression estimates (SV-adjusted, log2-normalized count data):
```{r}
#read in variance-transformed normalized count estimates for non-genotypic expression (genotypic effect and SVs regressed out)
nongeno_estimates <- readRDS(file=paste0(dir2, "output/nongenotypicLFCestimates.RDS"))

#spot check to confirm approximate normal distribution of log2 count data: looks good
#hist(nongeno_estimates[15,])

#transpose so genotypes in rows and genes in columns
nongenocounts_t <- t(nongeno_estimates)

#check variance across samples
gene_vars <- apply(nongenocounts_t, 2, var)
hist(gene_vars, breaks=100, main="distribution of gene variance")
# confirm all -Inf are removed after adding a small positive value to the zero counts in residuals calculation
sum(is.infinite(gene_vars)) == 0

#check for outliers/missing data with WGCNA function:
gsg <-goodSamplesGenes(nongenocounts_t, verbose=3)
gsg$allOK
#looks good
```

Pick power for correlations
```{r, eval=FALSE}
# Choose a set of soft-thresholding powers
powers = c(1:20) #seq(from = 1, to = 40, by = 5)

allowWGCNAThreads() 
# Call the network topology analysis function
sft = pickSoftThreshold(
  nongenocounts_t,   # <= Input data
  powerVector = powers,
  verbose = 5,
  networkType = "unsigned",
  corFnc="bicor")

saveRDS(sft, file=paste0(dir2, "output/picksoftthres_nongeno.RDS"))
```

Plot power threshold graphs
```{r}
sft <- readRDS(paste0(dir2, "output/picksoftthres_nongeno.RDS"))

par(mfrow = c(1,2)); cex1 = 0.9

#png(file=paste0(dir2, "figures/nongenonetwork_softthresholdpower_R2.png"))
  plot(sft$fitIndices[, 1],
       -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       xlab = "Soft Threshold (power)",
       ylab = "Scale Free Topology Model Fit, signed R^2",
       main = paste("Scale independence")
  )
  text(sft$fitIndices[, 1],
       -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       labels = powers, cex = cex1, col = "red"
  )
  abline(h = 0.8, col = "red")
  abline(h = 0.9, col = "red")
#dev.off()


#png(file=paste0(dir2, "figures/nongenonetwork_softthresholdpower_connectivity.png"))
plot(sft$fitIndices[, 1],
       sft$fitIndices[, 5],
       xlab = "Soft Threshold (power)",
       ylab = "Mean Connectivity",
       type = "n",
       main = paste("Mean connectivity")
  )
  text(sft$fitIndices[, 1],
       sft$fitIndices[, 5],
       labels = powers,
       cex = cex1, col = "red")
#dev.off()
```

Construct non-genetic network with WGCNA
```{r}
picked_power <- 7
temp_cor <- cor       
cor <- WGCNA::cor # Force it to use WGCNA cor function (fix a namespace conflict issue)
nongeno_netwk <- blockwiseModules(nongenocounts_t,                
                          
                          # == Adjacency Function ==
                          power = picked_power,       
                          networkType = "unsigned",   #includes both positive and negative correlations
                          corType = "bicor",  #more robust to outliers than pearson correlation
                          nThreads = 38,

                          # == Tree and Block Options ==
                          randomSeed = 14850,
                          deepSplit = 3, #value between 0 and 4, sensitivity to module splitting
                          pamRespectsDendro = F,
                          minModuleSize = 30,
                          maxBlockSize = ncol(nongenocounts_t), #allow for running in 1 block since we have enough computational power

                          # == Module Adjustments ==
                          reassignThreshold = 0.05,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)            
                          saveTOMs = T,
                          saveTOMFileBase = paste0("/workdir/ssv42/RNAseq_yr2/output/nongeno_ntwk_pwr", 
                          picked_power, "_"),

                          # == Output Options
                          numericLabels = T,
                          verbose = 2)

cor <- temp_cor # Return original cor function 
```

Look at non-genetic network modules:
```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(nongeno_netwk$colors)

blockselection <- 1 

# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  nongeno_netwk$dendrograms[[blockselection]],
  mergedColors[nongeno_netwk$blockGenes[[blockselection]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )

ggsave(file=paste0(dir2, "figures/nongeno_network_dendrogram_block", blockselection, ".png"))


nongeno_module_df <- data.frame(
  gene_id = names(nongeno_netwk$colors),
  colors = labels2colors(nongeno_netwk$colors)
)

#Get TOM
load(paste0(nongeno_netwk$TOMFiles))
nongeno_TOM <- as.matrix(TOM)


#export nongenotypic module dataframe
write_delim(nongeno_module_df,
            file = paste0(dir2, "output/nongeno_network_gene_modules.txt"),
            delim = "\t")
```

Test associations between modules and phenotypes
```{r}
# Get Module Eigengenes per cluster
MEs0 <- moduleEigengenes(nongenocounts_t, mergedColors)$eigengenes

# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0);
module_order = names(MEs0) %>% gsub("ME", "", .)

# Read in phenotypes
metadata_phenos_scaled <- readRDS(file=paste0(dir2, "output/scaled_pheno_data.RDS"))
#confirm sample order is the same: 
all(metadata_phenos_scaled$fullID == rownames(MEs0))
#combine with module eigengenes 
MEs0_wphenos <- cbind(MEs0,  
    (metadata_phenos_scaled %>% dplyr::select(Phytoene:Zeta.Carotene, Violaxanthin.peak.1:Lutein,
                        total_carotenoids:dry_matter, i_o_diff_TC))  ) 

# calculate correlations
MEs0_wphenos_cormat <- cor(MEs0_wphenos, use="pairwise.complete.obs")
MEs0_wphenos_cormat <- as.data.frame(MEs0_wphenos_cormat) 
MEs0_wphenos_cormat$module <- rownames(MEs0_wphenos_cormat)
MEs_wphenos_cormat <- MEs0_wphenos_cormat %>% pivot_longer(-module, names_to="phenotype") %>%
  filter(!str_detect(phenotype, "ME")) %>% 
  filter(str_detect(module, "ME")) %>%
  mutate(module = gsub("ME","", module)) %>%
  mutate(phenotype = factor(phenotype, levels=c("Phytoene", "Phytofluene", "Zeta.Carotene", "Beta.Carotene", "Lutein", 
                                     "Violaxanthin.peak.1", "Violaxanthin.peak.2", "total_carotenoids",
                                    "totalpre", "total_violaxanthin", "ratio_bc_v", "i_o_diff_TC", "dry_matter")) )


ggplot(MEs_wphenos_cormat, aes(y=phenotype, x=module, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Module-trait Relationships from Non-genetic Network", y = "Modules", fill="corr")

ggsave(filename=paste0(dir2, "figures/nongenontwk_module_trait_correlations.png"),
       height=6, width=18, units="in")

```

Merge network results with gene annotations
```{r}
#merge annotations with non-genotypic network module dataframe
nongeno_modules_annot <- nongeno_module_df %>% 
  mutate(locusName = gsub(".v8.1", "", gene_id)) %>%
  left_join(., fullgene_annotations)

#check how many genes in each class
table(nongeno_modules_annot$class)

#check which modules carotenoid genes are in
table(nongeno_modules_annot %>% filter(class=="carot") %>% dplyr::select(colors) )

#check carotenoid superpathway genes specifically
table(nongeno_modules_annot %>% filter(grepl("CAROTENOID-PWY", Pathways)) %>% dplyr::select(colors))
nongeno_modules_annot %>% filter(grepl("CAROTENOID-PWY", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

#check which modules carbohydrate genes are in
table(nongeno_modules_annot %>% filter(class=="carb") %>% dplyr::select(colors) )

#check starch synthesis pathway genes specifically
table(nongeno_modules_annot %>% filter(grepl("PWY-622", Pathways)) %>% dplyr::select(colors) )
nongeno_modules_annot %>% filter(grepl("PWY-622", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

#check starch degradation pathway genes specifically
table(nongeno_modules_annot %>% filter(grepl("PWY-6724", Pathways)) %>% dplyr::select(colors) )
nongeno_modules_annot %>% filter(grepl("PWY-6724", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

```

Select modules of interest from non-genotypic network and export for Cytoscape
```{r}
#check module sizes
table(nongeno_module_df$colors)

#identify modules most associated with the phenotypes
top_associations <- MEs_wphenos_cormat %>% group_by(phenotype) %>%
  slice_max(abs(value), n=3) 

top_modules <- unique((top_associations %>% filter(abs(value) > 0.4))$module)
print(top_modules)

nongeno_modules_of_interest <- "blue" #OR top_modules

# Pull out list of genes in that module
submod_nongeno <- nongeno_module_df %>%
  subset(colors %in% nongeno_modules_of_interest)

row.names(nongeno_module_df) <- nongeno_module_df$gene_id

# Subset TOM matrix for modules of interest
colnames(nongeno_TOM) <- nongeno_module_df$gene_id
rownames(nongeno_TOM) <- nongeno_module_df$gene_id

subset_nongenoTOM <- nongeno_TOM[rownames(nongeno_TOM) %in% submod_nongeno$gene_id,
                                colnames(nongeno_TOM) %in% submod_nongeno$gene_id]

# check dimensions
dim(subset_nongenoTOM)
format(object.size(subset_nongenoTOM), units="GB")

#export subset for Cytoscape
Cyexport <- exportNetworkToCytoscape(subset_nongenoTOM,
                         edgeFile = paste0(dir2, "output/nongeno_ntwk_modulesofinterest_CytoscapeInput-edges_blue.txt"),
                         nodeFile = paste0(dir2, "output/nongeno_ntwk_modulesofinterest_CytoscapeInput-nodes_blue.txt"),
                         weighted = TRUE, threshold = 0.1,
                         nodeNames = (colnames(subset_nongenoTOM) %>% gsub(".v8.1","",.)) ) #remove version tag from gene name

#export gene annotations for Cytoscape
subset_nongenoTOM_annot <- nongeno_modules_annot %>% 
  filter(gene_id %in% unique(c(rownames(subset_nongenoTOM), colnames(subset_nongenoTOM)))) %>%
  dplyr::select(gene_id, locusName, Description, colors, class) %>%
  mutate(Description = gsub(",",";",Description)) #remove commas to not confuse cytoscape reading csv

write.csv(subset_nongenoTOM_annot, file= paste0(dir2, "output/nongeno_ntwk_modulesofinterest_CytoscapeInput-nodes_annotations_blue.csv"), row.names = F, quote=F)
```

Create subnetwork of only carotenoid and carbohydrate genes
```{r}
car_gene_subset <- nongeno_modules_annot %>% filter(!is.na(class)) %>%
  mutate(Description = gsub(",",";",Description))

# Subset TOM matrix for genes of interest
nongenoTOM_car_genes <- nongeno_TOM[rownames(nongeno_TOM) %in% car_gene_subset$gene_id,
                                colnames(nongeno_TOM) %in% car_gene_subset$gene_id]

#export subset for Cytoscape
Cyexport <- exportNetworkToCytoscape(nongenoTOM_car_genes,
                         edgeFile = paste0(dir2, "output/nongeno_ntwk_modulesofinterest_CytoscapeInput-edges_cargenes.txt"),
                         nodeFile = paste0(dir2, "output/nongeno_ntwk_modulesofinterest_CytoscapeInput-nodes_cargenes.txt"),
                         weighted = TRUE, threshold = 0.01,
                         nodeNames = (colnames(nongenoTOM_car_genes) %>% gsub(".v8.1","",.)) ) #remove version tag from gene name

write.csv(car_gene_subset, file=paste0(dir2, "output/nongeno_ntwk_modulesofinterest_CytoscapeInput_cargene_annotations.csv"))
```

Calculate correlation matrix separately on genes of interest subset to preserve sign/direction information
```{r}
nongeno_subset_bicor <- bicor(
  nongenocounts_t[,colnames(nongenocounts_t) %in% car_gene_subset$gene_id] ) %>% 
  reshape2::melt() 

colnames(nongeno_subset_bicor) <- c("sourceNode", "targetNode", "bicor")

nongeno_subset_bicor_Cyto <- nongeno_subset_bicor %>%
  mutate(across(sourceNode:targetNode, ~ gsub(".v8.1", "", .))) %>%
  mutate(shared_name = paste0(sourceNode, " (interacts with) ", targetNode)) %>%
  mutate(cor_direction = case_when(bicor > 0 ~ "+", bicor <= 0 ~ "-"))

write.csv(nongeno_subset_bicor_Cyto, 
          file=paste0(dir2, "output/nongeno_netwk_cargene-subset_bicorrelations.csv"), 
          row.names = F)
```








# Construct genetic network
Read in genotypic expression estimates (SV-adjusted, log2-normalized count data):
```{r}
#read in variance-transformed normalized count estimates for genetic effect on expression (non-genetic effect and SVs regressed out)
geno_estimates <- readRDS(file=paste0(dir2, "output/genotypicLFCestimates.RDS"))

#spot check to confirm approximate normal distribution of log2 count data: looks good
hist(geno_estimates[10,])

#transpose so genotypes in rows and genes in columns
genocounts_t <- t(geno_estimates)

#check variance across samples
gene_vars <- apply(genocounts_t, 2, var)
hist(gene_vars, breaks=100, main="distribution of variance in genetic effect on gene expression")
# confirm all -Inf are removed after adding a small positive value to the zero counts in residuals calculation
sum(is.infinite(gene_vars)) == 0

#check for outliers/missing data with WGCNA function:
gsg <-goodSamplesGenes(genocounts_t, verbose=3)
gsg$allOK
#looks good
```

Pick power for correlations
```{r, eval=FALSE}
# Choose a set of soft-thresholding powers
powers = c(1:20) #seq(from = 1, to = 40, by = 5)

allowWGCNAThreads() 
# Call the network topology analysis function
sft = pickSoftThreshold(
  genocounts_t,   # <= Input data
  powerVector = powers,
  verbose = 5,
  networkType = "unsigned",
  corFnc="bicor")

saveRDS(sft, file=paste0(dir2, "output/picksoftthres_geno.RDS"))
```

Plot power threshold graphs
```{r}
sft <- readRDS(paste0(dir2, "output/picksoftthres_geno.RDS"))

par(mfrow = c(1,2)); cex1 = 0.9

#png(file=paste0(dir2, "figures/geno_network_softthresholdpower_R2.png"))
  plot(sft$fitIndices[, 1],
       -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       xlab = "Soft Threshold (power)",
       ylab = "Scale Free Topology Model Fit, signed R^2",
       main = paste("Scale independence")
  )
  text(sft$fitIndices[, 1],
       -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
       labels = powers, cex = cex1, col = "red"
  )
  abline(h = 0.8, col = "red")
  abline(h = 0.9, col = "red")
#dev.off()


#png(file=paste0(dir2, "figures/geno_network_softthresholdpower_connectivity.png"))
plot(sft$fitIndices[, 1],
       sft$fitIndices[, 5],
       xlab = "Soft Threshold (power)",
       ylab = "Mean Connectivity",
       type = "n",
       main = paste("Mean connectivity")
  )
  text(sft$fitIndices[, 1],
       sft$fitIndices[, 5],
       labels = powers,
       cex = cex1, col = "red")
#dev.off()
```

Construct genetic network with WGCNA
```{r}
picked_power <- 6
temp_cor <- cor       
cor <- WGCNA::cor # Force it to use WGCNA cor function (fix a namespace conflict issue)
geno_netwk <- blockwiseModules(genocounts_t,                
                          
                          # == Adjacency Function ==
                          power = picked_power,       
                          networkType = "unsigned",   #includes both positive and negative correlations
                          corType = "bicor",  #more robust to outliers than pearson correlation
                          nThreads = 38,

                          # == Tree and Block Options ==
                          randomSeed = 14850,
                          deepSplit = 2, #value between 0 and 4, sensitivity to module splitting
                          pamRespectsDendro = F,
                          minModuleSize = 30,
                          maxBlockSize = ncol(genocounts_t), #allow for running in 1 block since we have enough computational power

                          # == Module Adjustments ==
                          reassignThreshold = 0.05,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)            
                          saveTOMs = T,
                          saveTOMFileBase = paste0("/workdir/ssv42/RNAseq_yr2/output/geno_ntwk_pwr", 
                          picked_power, "_ds2_"),

                          # == Output Options
                          numericLabels = T,
                          verbose = 2)

cor <- temp_cor # Return original cor function 

saveRDS(geno_netwk, file=paste0(dir2, "output/geno_netwk_ds2.RDS"))
```

Look at genetic network modules:
```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(geno_netwk$colors)

blockselection <- 1 

# Plot the dendrogram and the module colors underneath
png(file=paste0(dir2, "figures/geno_network_ds2_dendrogram_block", blockselection, ".png"),
    width=8, height=5, units="in", res=400)
plotDendroAndColors(
  geno_netwk$dendrograms[[blockselection]],
  mergedColors[geno_netwk$blockGenes[[blockselection]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
dev.off()

#print(geno_dendro)
#geno_dendro


geno_module_df <- data.frame(
  gene_id = names(geno_netwk$colors),
  colors = labels2colors(geno_netwk$colors)
)

#Get TOM
load(paste0(geno_netwk$TOMFiles))
geno_TOM <- as.matrix(TOM)


#export genotypic module dataframe
write_delim(geno_module_df,
            file = paste0(dir2, "output/geno_network_ds2_gene_modules.txt"),
            delim = "\t")
```



Test associations between genetic modules and phenotypes
```{r}
# Get Module Eigengenes per cluster
MEs0_g <- moduleEigengenes(genocounts_t, mergedColors)$eigengenes

# Reorder modules so similar modules are next to each other
MEs_genetic <- orderMEs(MEs0_g);
g_module_order = names(MEs0_g) %>% gsub("ME", "", .)

# Read in phenotype BLUEs
allpheno_BLUEs_int <- readRDS(file=paste0(dir2, "output/PhenotypeBLUEs.RDS"))
phenoBLUEs <- t(allpheno_BLUEs_int)

#confirm sample order is the same (ignoring underscore formatting differences): 
all(gsub("_","",rownames(phenoBLUEs)) == gsub("_","",rownames(MEs_genetic)))

#combine with module eigengenes
MEs_genetic_wphenos <- cbind(MEs_genetic,  phenoBLUEs) 

# calculate correlations
MEs_genetic_wphenos_cormat <- cor(MEs_genetic_wphenos, use="pairwise.complete.obs")
MEs_genetic_wphenos_cormat <- as.data.frame(MEs_genetic_wphenos_cormat) 
MEs_genetic_wphenos_cormat$module <- rownames(MEs_genetic_wphenos_cormat)
MEs_genetic_wphenos_df <- MEs_genetic_wphenos_cormat %>% pivot_longer(-module, names_to="phenotype") %>%
  filter(!str_detect(phenotype, "ME")) %>% 
  filter(str_detect(module, "ME")) %>%
  mutate(module = gsub("ME","", module)) %>%
  mutate(phenotype = factor(phenotype, levels=c("Phytoene", "Phytofluene", "Zeta.Carotene", "Beta.Carotene",
                                                "Lutein", "total_violaxanthin", "total_carotenoids", 
                                                "i_o_diff_TC", "ratio_cis_bc", "ratio_viola_bc", 
                                                "dry_matter", "i_o_diff_DM")) ) %>%
  filter(!is.na(phenotype))

ggplot(MEs_genetic_wphenos_df, aes(y=phenotype, x=module, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Module-trait Relationships from Genetic Network", 
       y = "Phenotype", x = "Gene expression module", fill="corr")

ggsave(filename=paste0(dir2, "figures/geno_ntwk_ds2_module_trait_correlations.png"),
       height=6, width=18, units="in")
```

Merge genetic network results with gene annotations
```{r}
#merge annotations with non-genotypic network module dataframe
geno_modules_annot <- geno_module_df %>% 
  mutate(locusName = gsub(".v8.1", "", gene_id)) %>%
  left_join(., fullgene_annotations)

#check how many genes in each class
table(geno_modules_annot$class)

#check which modules carotenoid genes are in
table(geno_modules_annot %>% filter(class=="carot") %>% dplyr::select(colors) )

#check carotenoid superpathway genes specifically
table(geno_modules_annot %>% filter(grepl("CAROTENOID-PWY", Pathways)) %>% dplyr::select(colors))
geno_modules_annot %>% filter(grepl("CAROTENOID-PWY", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

#check which modules carbohydrate genes are in
table(geno_modules_annot %>% filter(class=="carb") %>% dplyr::select(colors) )

#check starch synthesis pathway genes specifically
table(geno_modules_annot %>% filter(grepl("PWY-622", Pathways)) %>% dplyr::select(colors) )
geno_modules_annot %>% filter(grepl("PWY-622", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

#check starch degradation pathway genes specifically
table(geno_modules_annot %>% filter(grepl("PWY-6724", Pathways)) %>% dplyr::select(colors) )
geno_modules_annot %>% filter(grepl("PWY-6724", Pathways)) %>% dplyr::select(colors, locusName, Description) %>%
  rmarkdown::paged_table()

```

Select modules of interest from genotypic network and export for Cytoscape
```{r}
#check module sizes
table(geno_module_df$colors)

#identify modules most associated with the phenotypes
geno_top_associations <- MEs_genetic_wphenos_df %>% group_by(phenotype) %>%
  slice_max(abs(value), n=1) 

geno_top_modules <- unique((geno_top_associations %>% filter(abs(value) > 0.5))$module)
print(geno_top_modules)

geno_modules_of_interest <- geno_top_modules #pick individual module or top_modules

# Pull out list of genes in that module
submod_geno <- geno_module_df %>%
  subset(colors %in% geno_modules_of_interest)

row.names(geno_module_df) <- geno_module_df$gene_id

# Subset TOM matrix for modules of interest
colnames(geno_TOM) <- geno_module_df$gene_id
rownames(geno_TOM) <- geno_module_df$gene_id

subset_genoTOM <- geno_TOM[rownames(geno_TOM) %in% submod_geno$gene_id,
                                colnames(geno_TOM) %in% submod_geno$gene_id]

# check dimensions
dim(subset_genoTOM)
format(object.size(subset_genoTOM), units="GB")

#export subset for Cytoscape
Cyexport <- exportNetworkToCytoscape(subset_genoTOM,
                         edgeFile = paste0(dir2, "output/geno_ntwk_ds2_modulesofinterest_CytoscapeInput-edges_", 
                                           geno_modules_of_interest[1], ".txt"),
                         nodeFile = paste0(dir2, "output/geno_ntwk_ds2_modulesofinterest_CytoscapeInput-nodes_", 
                                           geno_modules_of_interest[1], ".txt"),
                         weighted = TRUE, threshold = 0.1,
                         nodeNames = (colnames(subset_genoTOM) %>% gsub(".v8.1","",.)) ) #remove version tag from gene name

#export gene annotations for Cytoscape
subset_genoTOM_annot <- geno_modules_annot %>% 
  filter(gene_id %in% unique(c(rownames(subset_genoTOM), colnames(subset_genoTOM)))) %>%
  dplyr::select(gene_id, locusName, Description, colors, class) %>%
  mutate(Description = gsub(",",";",Description)) #remove commas to not confuse cytoscape reading csv

write.csv(subset_genoTOM_annot, file= paste0(dir2, "output/geno_ntwk_ds2_modulesofinterest_CytoscapeInput-nodes_annotations_",
                                             geno_modules_of_interest[1], ".csv"), row.names = F, quote=F)




#Calculate correlation matrix separately on genes of interest subset to preserve sign/direction information

geno_subset_bicor <- bicor(
  genocounts_t[,colnames(genocounts_t) %in% submod_geno$gene_id] ) %>% 
  reshape2::melt() 

colnames(geno_subset_bicor) <- c("sourceNode", "targetNode", "bicor")

geno_subset_bicor_Cyto <- geno_subset_bicor %>%
  mutate(across(sourceNode:targetNode, ~ gsub(".v8.1", "", .))) %>%
  mutate(shared_name = paste0(sourceNode, " (interacts with) ", targetNode)) %>%
  mutate(cor_direction = case_when(bicor > 0 ~ "+", bicor <= 0 ~ "-"))

write.csv(geno_subset_bicor_Cyto, 
          file=paste0(dir2, "output/geno_ntwk_ds2_modulesofinterest__bicorrelations.csv"), 
          row.names = F)
```

Create subnetwork of only carotenoid and carbohydrate genes
```{r}
car_gene_subset <- geno_modules_annot %>% filter(!is.na(class)) %>%
  mutate(Description = gsub(",",";",Description))

# Subset TOM matrix for genes of interest
genoTOM_car_genes <- geno_TOM[rownames(geno_TOM) %in% car_gene_subset$gene_id,
                                colnames(geno_TOM) %in% car_gene_subset$gene_id]

#export subset for Cytoscape
Cyexport <- exportNetworkToCytoscape(genoTOM_car_genes,
                         edgeFile = paste0(dir2, "output/geno_ntwk_ds2_modulesofinterest_CytoscapeInput-edges_cargenes.txt"),
                         nodeFile = paste0(dir2, "output/geno_ntwk_ds2_modulesofinterest_CytoscapeInput-nodes_cargenes.txt"),
                         weighted = TRUE, threshold = 0.01,
                         nodeNames = (colnames(genoTOM_car_genes) %>% gsub(".v8.1","",.)) ) #remove version tag from gene name

#export gene annotations for subset
write.csv(car_gene_subset, file=paste0(dir2, "output/geno_ntwk_ds2_modulesofinterest_CytoscapeInput_cargene_annotations.csv"))



#Calculate correlation matrix separately on genes of interest subset to preserve sign/direction information
geno_subset_bicor <- bicor(
  genocounts_t[,colnames(genocounts_t) %in% car_gene_subset$gene_id] ) %>% 
  reshape2::melt() 

colnames(geno_subset_bicor) <- c("sourceNode", "targetNode", "bicor")

geno_subset_bicor_Cyto <- geno_subset_bicor %>%
  mutate(across(sourceNode:targetNode, ~ gsub(".v8.1", "", .))) %>%
  mutate(shared_name = paste0(sourceNode, " (interacts with) ", targetNode)) %>%
  mutate(cor_direction = case_when(bicor > 0 ~ "+", bicor <= 0 ~ "-"))

write.csv(geno_subset_bicor_Cyto, 
          file=paste0(dir2, "output/geno_netwk_ds2_cargene-subset_bicorrelations.csv"), 
          row.names = F)
```


Test how often carotenoid and carbohydrate genes cluster in the same modules:
```{r}
#shuffle the group of carotenoid genes in one module to a new module, randomly drawn with probability based on the module size compared to total size (keeping carotenoid genes clustered with each other together)

# count how many times they cluster with carbohydrate genes (even 1 gene = yes? or count how many carb genes are in that module?)
#geno_modules_annot
```


Identify hub genes:
```{r}
#double check gene order is the same
all(colnames(genocounts_t) == geno_module_df$gene_id)

geno_tophubs <- chooseTopHubInEachModule(genocounts_t, 
                                         colorh=geno_module_df$colors,
                                         power = 6, type="unsigned")

geno_tophubs_df <- data.frame("hub_gene" = geno_tophubs)
geno_tophubs_df$module <- rownames(geno_tophubs_df) 
geno_tophubs_annot <- geno_tophubs_df %>% left_join(geno_modules_annot, by=c("hub_gene" = "gene_id"))

#check the hub genes of the major carotenoid/carbohydrate modules
#hubstocheck <- names(tail(sort(table(car_gene_subset$colors), descending=T), n=10))

hubstocheck <- geno_top_modules
geno_hubs_topmodules <- geno_tophubs_annot %>% filter(module %in% hubstocheck)


#hubGeneSignificance(datKME, GS)
```

Network screening for genes associated with traits:
```{r}
geno_netscreen <- automaticNetworkScreening(
  genocounts_t,
  y = 
  power = picked_power,
  networkType = "unsigned",
  detectCutHeight = 0.995,
  minModuleSize = min(20, ncol(as.matrix(datExpr))/2),
  datME = NULL)
```

Test TOMplot:
```{r}
TOMplot(dissim=geno_TOM, dendro=geno_netwk$dendrograms[[1]], Colors=geno_module_df$colors)
```













# Next analysis step recommendations from Elise:
Kohonen package

string database of interactions with transcriptomic data in Cytoscape

Scatterplot of gene significance (GS) vs module membership (MM)

Pathway enrichment of trait-related modules:
enricher package
enrichplot
ontologyIndex 



#OLD
## Genetic network
Pick a soft threshold power near the curve of the plot.
Choose lowest possible power term where topology approximately fits a scale free network (on or above red horizontal line).
Run WGCNA on genetic LFCs:
```{r,eval=F}
picked_power <- 9
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk <- blockwiseModules(ncounts_t,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "unsigned",   #includes both positive and negative correlations

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,

                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = T,
                          verbose = 3)

cor <- temp_cor # Return original cor function 
```

Look at the modules
```{r,eval=F}
# Convert labels to colors for plotting
mergedColors = labels2colors(netwk$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  netwk$dendrograms[[1]],
  mergedColors[netwk$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )


module_df <- data.frame(
  gene_id = names(netwk$colors),
  colors = labels2colors(netwk$colors)
)

module_df[1:5,]
#>            gene_id    colors
#> 1 AC149818.2_FG001      blue
#> 2 AC149829.2_FG003      blue
#> 3 AC182617.3_FG001      blue
#> 4 AC186512.3_FG001 turquoise
#> 5 AC186512.3_FG007 turquoise

write_delim(module_df,
            file = "BLUE_gene_modules.txt",
            delim = "\t")
```

Calculate phenotype BLUEs to correlate with eigengene BLUEs
```{r, eval=F}
library(lme4)

metadata_phenos_scaled <- readRDS(file=paste0(dir2, "output/scaled_pheno_data.RDS"))

#Carotenoid BLUEs
carot_data <- metadata_phenos_scaled %>% dplyr::select(fullID:harvest_date, carot_extraction_date, 
                                                       Phytoene:Zeta.Carotene, Violaxanthin.peak.1:Lutein,
                                                       total_carotenoids:total_violaxanthin) %>%
  mutate(root = as.factor(root), block = as.factor(block)) %>%
  mutate(ratio_upstream_total = ( sum(Phytoene, Phytofluene, Zeta.Carotene) / total_carotenoids), 
         ratio_bc_viola = Beta.Carotene / total_violaxanthin) 

#pull out carotenoid compound/trait names
carot_traits <- colnames(carot_data)[13:ncol(carot_data)]

#initiate BLUE storage
carot_BLUEs <- matrix(nrow = length(carot_traits), ncol=length(unique(metadata_phenos_scaled$genotype)))

#loop through each trait to calculate BLUEs
for(i in seq_along(carot_traits)){
  
  traiti <- carot_traits[i]
  print(traiti)
  
  traiti_mod <- lm( formula( paste0(traiti, "~ 1 + genotype + year + timept + layer + block + carot_extraction_date") ),  
                      data = carot_data)
  
  #extract genotypic coefficients and save to matrix
  traiti_BLUEs <- traiti_mod$coefficients[c(1, grep("genotype", names( traiti_mod$coefficients)))] 
  carot_BLUEs[i,] <- traiti_BLUEs
}
carot_BLUEs <- as.data.frame(carot_BLUEs) 
colnames(carot_BLUEs) <- names(traiti_BLUEs)
rownames(carot_BLUEs) <- carot_traits


#Calculate dry matter BLUEs:
dm_mod <- lm( dry_matter ~ 1 + genotype + year + timept + layer + block,  
                      data = metadata_phenos_scaled)

dm_BLUEs <- dm_mod$coefficients[c(1, grep("genotype", names( dm_mod$coefficients )))] 


#Merge with carotenoid BLUEs:
allpheno_BLUEs <- rbind(carot_BLUEs, dm_BLUEs)
rownames(allpheno_BLUEs)[nrow(allpheno_BLUEs)] <- "dry_matter"


#add intercept to every column
allpheno_BLUEs_int <- allpheno_BLUEs %>% mutate(across(-1, ~ .x + allpheno_BLUEs$`(Intercept)`))
colnames(allpheno_BLUEs_int)[1] <- "genotypeA03"

#save phenotype BLUEs
saveRDS(allpheno_BLUEs_int, file=paste0(dir2, "output/PhenotypeBLUEs.RDS"))



```

Correlate module eigengenes with phenotypes
```{r, eval=F}
# Get Module Eigengenes per cluster
MEs0 <- moduleEigengenes(ncounts_t, mergedColors)$eigengenes

# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0);
#colnames(MEs0) <- gsub("ME", "", colnames(MEs0))
module_order = names(MEs0) %>% gsub("ME", "", .)

# Add phenotypes
#check order is the same: rownames(phenoBLUEs); rownames(MEs0)
phenoBLUEs <- t(allpheno_BLUEs_int) #transpose first
MEs0_wphenos <- cbind(MEs0,  phenoBLUEs) 

# calculate correlations
MEs0_wphenos_cormat <- cor(MEs0_wphenos)
MEs0_wphenos_cormat <- as.data.frame(MEs0_wphenos_cormat) 
MEs0_wphenos_cormat$module <- rownames(MEs0_wphenos_cormat)
MEs_wphenos_cormat <- MEs0_wphenos_cormat %>% pivot_longer(-module, names_to="phenotype") %>%
  filter(!str_detect(phenotype, "ME")) %>% 
  filter(str_detect(module, "ME")) %>%
  mutate(module = gsub("ME","", module)) %>%
  mutate(phenotype = factor(phenotype, levels=c("Phytoene", "Phytofluene", "Zeta.Carotene", "Beta.Carotene", "Lutein", 
                                                     "Violaxanthin.peak.1", "Violaxanthin.peak.2", "total_violaxanthin", 
                                                    "ratio_upstream_total", "ratio_bc_viola",  "total_carotenoids", "dry_matter")))


ggplot(MEs_wphenos_cormat, aes(y=phenotype, x=module, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")

ggsave(filename=paste0(dir2, "figures/genetic_modules_trait_correlations.png"),
       height=6, width=18, units="in")

#plot only strong correlations
MEs_wphenos_strongcor <- MEs_wphenos_cormat %>% filter(abs(value) >= 0.4)

ggplot(MEs_wphenos_strongcor, aes(y=phenotype, x=module, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Strongest Module-trait Relationships", y = "Modules", fill="corr")

ggsave(filename=paste0(dir2, "figures/genetic_modules_trait_strongest_correlations.png"),
        height=6, width=18, units="in")
```

Examine expression profiles across genotypes for modules of interest
```{r, eval=F}
# pick out modules of interest here --  #correlated with total carotenoids and/or DM
gen_modules_of_interest = c("blue", "darkslateblue", "white", "salmon", "grey",
                        "lightpink4", "violet", "darkred")

# Pull out list of genes in that module
submod = module_df %>%
  subset(colors %in% modules_of_interest)

row.names(module_df) = module_df$gene_id
subexpr = genoLFCs[submod$gene_id,]

submod_df = data.frame(subexpr) %>%
  mutate(
    gene_id = row.names(.)
  ) %>%
  pivot_longer(-gene_id) %>%
  mutate(
    module = module_df[gene_id,]$colors
  )

submod_df %>% ggplot(., aes(x=name, y=value, group=gene_id)) +
  geom_line(aes(color = module),
            alpha = 0.2) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  facet_grid(rows = vars(module)) +
  labs(x = "treatment",
       y = "normalized expression")


```

Export network to cytoscape
```{r, eval=F}
genes_of_interest = module_df %>%
  subset(colors %in% modules_of_interest)

expr_of_interest = genoLFCs[genes_of_interest$gene_id,]

TOM = TOMsimilarityFromExpr(t(expr_of_interest),
                            power = picked_power)

row.names(TOM) = row.names(expr_of_interest)
colnames(TOM) = row.names(expr_of_interest)

edge_list = data.frame(TOM) %>%
  mutate(
    gene1 = row.names(.)
  ) %>%
  pivot_longer(-gene1) %>%
  dplyr::rename(gene2 = name, correlation = value) %>%
  unique() %>%
  subset(!(gene1==gene2)) %>%
  mutate(
    module1 = module_df[gene1,]$colors,
    module2 = module_df[gene2,]$colors
  )

write_delim(edge_list,
            file = "genoLFC_modulesofinterest_edgelist.tsv",
            delim = "\t")

```



