---
title: "carotenoid_extractions"
output: html_document
date: "2024-07-19"
---
Loading R packages
```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggplot2)

```


# Read in data
```{r}
carotenoids <- read.csv("../data/Cassava_Carotenoid_Extraction.csv") %>% #uplc/uv-vis data
  filter(tube_ID != 168) %>% #exclude a duplicate with outlier total carotenoids
  dplyr::select(-c(X:X.3)) #remove empty columns 

weights <- read.csv("../data/carotenoid_sample_weights.csv") #sample weights for carotenoid extractions

dry_matter <- read.csv("../data/3_harvest_data.csv") %>% #dry matter and metadata
  mutate(tube_ID = as.character(samplenum)) %>% dplyr::select(-samplenum) %>%
  mutate(block = substr(as.character(plant), 1, 1), #get block number from first digit of plant number
         timept = case_when(harvest_date %in% c("5/29/23", "5/30/23") ~ "6MAP", 
                            harvest_date %in% c("6/20/23", "6/21/23") ~ "7MAP", 
                            harvest_date %in% c("7/6/23") ~ "8MAP", 
                            harvest_date %in% c("7/28/23") ~ "9MAP", T ~ NA) )
```


# Data curation
Change variable types, making DM above 100 be NA, standarizing with sample weights and changing units from ng/g to ug/g.

```{r}
joined_df <- full_join(x=carotenoids, y=weights, by= c("tube_ID" = "tube_number")) %>%  
  left_join(y= dry_matter, by=c("tube_ID", "sample_ID" = "old_sample_ID")) %>%
  dplyr::select(-sample_ID) %>% #drop uncorrected sample ID column #tubeID %>%
  rename(originaltubelabel = tubeID) %>% 
  mutate(Extr.Set.Date = as.factor(Extr.Set.Date), 
         sample_weight_g = as.numeric(sample_weight_g), 
         block = as.factor(block), 
         plant = as.factor(plant), 
         root = as.factor(root), across(Phytoene:Lutein, as.numeric)) %>%
  mutate(dry_matter = case_when(`dry_matter._complete` > 100 ~ NA, TRUE ~ `dry_matter._complete`)) #filter out impossible percentages

joined_df <- joined_df %>% 
  mutate(across(Phytoene:Lutein, function(x){x/joined_df$sample_weight_g})) %>%  #standardize carotenoids by starting weight
  mutate(across(Phytoene:Lutein, function(x){x*0.001})) %>% #convert ng/g to ug/g
  mutate(across(Phytoene:Lutein, ~ case_when(tube_ID == "2" ~ NA, T ~ . )) ) %>% #exclude tube 02 spilled during dry down
  mutate(total_carotenoids = rowSums(dplyr::select(., Phytoene:Lutein)), #sum total carotenoids
         total_violaxanthin = Violaxanthin.peak.1 + Violaxanthin.peak.2) %>% #sum total violaxanthin
  mutate(Beta.Carotene = case_when(Beta.Carotene == 0 ~ NA, TRUE ~ Beta.Carotene)) %>% #filter out impossible percentages
  mutate(genotype.category = case_when(
    startsWith(genotypeID, "A") ~ "African",
    startsWith(genotypeID, "C") ~ "Latin American",
    startsWith(genotypeID, "X") ~ "Hybrid")) %>%
  mutate(genotypeID = case_when(grepl("S", tube_ID) == T ~ "standard", T ~ genotypeID)) %>%
  dplyr::select(-c(dry_matter., dry_matter._complete))


#write.csv(joined_df, "../data/CarotenoidsHPLCdata_merged_yr2.csv")
joined_df = read.csv("../data/CarotenoidsHPLCdata_merged_yr2.csv")
```




Detecting and remove outliers 

```{r}
library(outliers)
#Grubbs test
outliertest <- grubbs.test(joined_df$dry_matter)
#outliertest

#removing outliers of DM
joined_df[joined_df$tube_ID == "228", "dry_matter" ] <- NA

outliertest2 <- grubbs.test(joined_df$dry_matter)
# #outliertest2
joined_df[joined_df$tube_ID == "62", "dry_matter" ] <- NA

outliertest3 <- grubbs.test(joined_df$total_carotenoids)
# # #outliertest3
joined_df[joined_df$tube_ID == "349", "total_carotenoids" ] <- NA

```

# Carotenoid composition
```{r}
data_long <- joined_df %>% mutate(CisCarotenes = Phytoene + Phytofluene + Zeta.Carotene) %>%
  pivot_longer(cols = c(Beta.Carotene:Lutein, CisCarotenes), names_to = "Carotenoid", values_to = "value") 

ggplot(data_long, aes(x = "", y = value, fill = Carotenoid)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  theme_void() +
  labs(title = "Carotenoid composition") 

```



# Check the technical consistency of the standards  

```{r}
standards <- joined_df %>% 
  filter(str_starts(tube_ID, "^S")) 

std_means <- standards %>% summarise(across(c(Phytoene:Lutein, total_carotenoids, total_violaxanthin), mean, na.rm = T)) #means
std_sd <- standards %>% summarise(across(c(Phytoene:Lutein, total_carotenoids, total_violaxanthin), sd, na.rm = T)) #standard deviation
std_cv <- (std_sd/std_means)*100 #coefficient of variation in % 

head(std_cv)

ggplot(data = standards)+
  aes(x = genotypeID, y = total_carotenoids)+
  geom_boxplot()
```

# Data Analysis 

Filtering into categories
```{r}
inner_df <- filter(joined_df, layer == "inner")
outer_df <- filter(joined_df, layer == "outer")

timept_1 <- filter(joined_df, timept == "6MAP")
timept_2 <- filter(joined_df, timept == "7MAP")
timept_3 <- filter(joined_df, timept == "8MAP")
timept_4 <- filter(joined_df, timept == "9MAP")
```


Phenotypic correlation tests 
```{r}
# phenotypic correlation between TC & DM 
TCDM_cor <- cor.test(joined_df$total_carotenoids, joined_df$dry_matter, method="kendall")
TCDM_cor


#TC/DM correlation between layers 
cor.test(inner_df$total_carotenoids, inner_df$dry_matter, method="kendall") 
cor.test(outer_df$total_carotenoids, outer_df$dry_matter, method="kendall")



#TC/DM phenotypic correlation across time points 
cor.test(timept_1$total_carotenoids, timept_1$dry_matter)

cor.test(timept_2$total_carotenoids, timept_2$dry_matter)
 
cor.test(timept_3$total_carotenoids, timept_3$dry_matter)

cor.test(timept_4$total_carotenoids, timept_4$dry_matter)

```

- TC and DM are significantly negatively phenotypically correlated, r=-0.24
- Inner layer: TC is negatively correlated with DM 
- Outer layer: TC is not significantly correlated with DM 
- TC and DM are most phenotypically negatively correlated at timepoint 4, -0.41


Examine genetic correlations with linear models
```{r}
library(lme4)

#overall genetic correlation and heritability
DMBLUPmodel <- lmer(data = joined_df, dry_matter ~ block + (1|genotypeID) + layer + timept + (1|plant))
TCBLUPmodel <- lmer(data = joined_df, total_carotenoids ~ block + (1|genotypeID) +  layer + timept + (1|plant))

overallgencor <- cor.test(ranef(DMBLUPmodel)[[1]][,1], ranef(TCBLUPmodel)[[1]][,1])

H2DM <- as.data.frame(summary(DMBLUPmodel)$varcor)[2,"vcov"] / sum(as.data.frame(summary(DMBLUPmodel)$varcor)[,"vcov"]*c((1/4), 1, (1/24)))
H2TC <- as.data.frame(summary(TCBLUPmodel)$varcor)[2,"vcov"] / sum(as.data.frame(summary(TCBLUPmodel)$varcor)[,"vcov"]*c((1/4), 1, (1/24)))

cat("Overall H2 of DM: ", round(H2DM,2), "\n")
cat("Overall H2 of TC: ", round(H2TC,2), "\n")


DMBLUPmodel_1 <- lmer(data = timept_1, dry_matter ~ block + (1|genotypeID) + layer)
H2DM_1 <- as.data.frame(summary(DMBLUPmodel_1)$varcor)[1,"vcov"] / sum(as.data.frame(summary(DMBLUPmodel_1)$varcor)[,"vcov"])
TCBLUPmodel_1 <- lmer(data = timept_1, total_carotenoids ~ block + (1|genotypeID) + layer)
H2TC_1 <-  as.data.frame(summary(TCBLUPmodel_1)$varcor)[1,"vcov"] / sum(as.data.frame(summary(TCBLUPmodel_1)$varcor)[,"vcov"])
tp1gencor <- cor.test(ranef(DMBLUPmodel_1)[[1]][,1], ranef(TCBLUPmodel_1)[[1]][,1])
tp1layer <- as.data.frame(summary(DMBLUPmodel_1)$coefficients)["layerouter", "Estimate"]

DMBLUPmodel_2 <- lmer(data = timept_2, dry_matter ~ block + (1|genotypeID) + layer)
H2DM_2 <- as.data.frame(summary(DMBLUPmodel_2)$varcor)[1,"vcov"] / sum(as.data.frame(summary(DMBLUPmodel_2)$varcor)[,"vcov"])
TCBLUPmodel_2 <- lmer(data = timept_2, total_carotenoids ~ block + (1|genotypeID) + layer)
H2TC_2 <-  as.data.frame(summary(TCBLUPmodel_2)$varcor)[1,"vcov"] / sum(as.data.frame(summary(TCBLUPmodel_2)$varcor)[,"vcov"])
tp2gencor <- cor.test(ranef(DMBLUPmodel_2)[[1]][,1], ranef(TCBLUPmodel_2)[[1]][,1])
tp2layer <- as.data.frame(summary(DMBLUPmodel_2)$coefficients)["layerouter", "Estimate"]



DMBLUPmodel_3 <- lmer(data = timept_3, dry_matter ~ block + (1|genotypeID) + layer)
H2DM_3 <- as.data.frame(summary(DMBLUPmodel_3)$varcor)[1,"vcov"] / sum(as.data.frame(summary(DMBLUPmodel_3)$varcor)[,"vcov"])
TCBLUPmodel_3 <- lmer(data = timept_3, total_carotenoids ~ block + (1|genotypeID) + layer)
H2TC_3 <-  as.data.frame(summary(TCBLUPmodel_3)$varcor)[1,"vcov"] / sum(as.data.frame(summary(TCBLUPmodel_3)$varcor)[,"vcov"])
tp3gencor <- cor.test(ranef(DMBLUPmodel_3)[[1]][,1], ranef(TCBLUPmodel_3)[[1]][,1])
tp3layer <- as.data.frame(summary(DMBLUPmodel_3)$coefficients)["layerouter", "Estimate"]


DMBLUPmodel_4 <- lmer(data = timept_4, dry_matter ~ block + (1|genotypeID) + layer)
H2DM_4 <- as.data.frame(summary(DMBLUPmodel_4)$varcor)[1,"vcov"] / sum(as.data.frame(summary(DMBLUPmodel_4)$varcor)[,"vcov"])
TCBLUPmodel_4 <- lmer(data = timept_4, total_carotenoids ~ block + (1|genotypeID) + layer)
H2TC_4 <-  as.data.frame(summary(TCBLUPmodel_4)$varcor)[1,"vcov"] / sum(as.data.frame(summary(TCBLUPmodel_4)$varcor)[,"vcov"])
tp4gencor <- cor.test(ranef(DMBLUPmodel_4)[[1]][,1], ranef(TCBLUPmodel_4)[[1]][,1])
tp4layer <- as.data.frame(summary(DMBLUPmodel_4)$coefficients)["layerouter", "Estimate"]


cat("genetic correlations across timepoints: " ,tp1gencor$estimate, tp2gencor$estimate, tp3gencor$estimate, tp4gencor$estimate, "\n")

cat("H2 of DM across timepoints: ", H2DM_1, H2DM_2, H2DM_3, H2DM_4, "\n")

cat("H2 of TC across timepoints: ", H2TC_1, H2TC_2, H2TC_3, H2TC_4, "\n")

cat("layer effect across timepoints: ", tp1layer, tp2layer, tp3layer, tp4layer)

```

- TC and DM are most genetically negatively correlated at timepoint 4, -0.60
- timepoint 4 also has the highest heritability for DM and the second highest for TC

  
  
#Pick 4 genotypes to focus on for RNAseq across all timepoints:
For the four most important traits (DM, TC, phytoene, beta-carotene), fit a line through the timepoints (one for inner, one for outer layer)
Get genotype/layer mean and std. error for each timepoint, divide by standard error (?)
Calculate the distance between each genotype/layer scaled mean and the line
Pick the 4 genotypes with the lowest total distance across timepoints and traits 
Check that the selected genotypes have complete TC/DM data and have tissue available for RNA. (nobs = 3)

To identify 4 genotypes representative of the pattern across time points for sampling gene expression across time points in the RNAseq experiment, the means of the three root replicates per plant/tissue layer were calculated and divided by the standard error. Linear models of DM and TC by time point were fit. The Euclidean distance between each standardized mean and the regression line was calculated and the genotypes were ranked by mean distance across tissue layers and time points. The 4 genotypes with the highest rank average between the DM and TC models, indicating that they most closely follow the overall trend of DM and TC across time points, were selected.
```{r}
plotdata <- joined_df %>% filter(!is.na(timept)) %>% 
  mutate(timept_numeric = as.numeric(gsub("MAP", "", timept)))  %>%
  filter(!is.na(total_carotenoids), !is.na(dry_matter)) %>%   
  filter(dry_matter < 60) %>%
  group_by(genotypeID, plant, timept_numeric, layer) %>%
  summarise(mean_TC = mean(total_carotenoids), se_TC = sd(total_carotenoids)/length(total_carotenoids),
            mean_phytoene = mean(Phytoene), se_phytoene = sd(Phytoene)/length(Phytoene),
            mean_BC = mean(Beta.Carotene), se_BC = sd(Beta.Carotene)/length(Beta.Carotene),
            mean_DM = mean(dry_matter), se_DM = sd(dry_matter)/length(dry_matter),
            nobs_TC = length(total_carotenoids), nobs_DM = length(dry_matter)) %>%  
  filter(nobs_TC == 3 & nobs_DM == 3) 

complete_obs_genos <- plotdata %>%
  filter(genotypeID %in% names(table(plotdata$genotypeID))[(table(plotdata$genotypeID) == 8)])



#[don't] divide means by standard error [but code to do it commented out]
plotdata_scaled <- plotdata %>%
  group_by(genotypeID, timept_numeric, layer) %>%
  filter(!is.na(se_DM) & !is.na(se_phytoene))  %>% # remove sample with only a single datapoint so SE is NA
  reframe(mean_TC_scaled = mean_TC, #/ se_TC,
            mean_phytoene_scaled = mean_phytoene, #/ se_phytoene,
            mean_BC_scaled = mean_BC, #/ se_BC,
            mean_DM_scaled = mean_DM) %>% # / se_DM) %>%
  filter(!is.na(mean_phytoene_scaled)) #remove observations where SE is 0 and can't be scaled 

TCscaledmod <- lm(mean_TC_scaled ~ timept_numeric*layer + genotypeID, data=plotdata_scaled)
fitted_TCscaled <- predict(TCscaledmod)
distance_TC <- sqrt( (plotdata_scaled$mean_TC_scaled - fitted_TCscaled) ^2)

phytoscaledmod <- lm(mean_phytoene_scaled ~ timept_numeric*layer + genotypeID, data=plotdata_scaled)
fitted_phytoscaled <- predict(phytoscaledmod)
distance_phytoene <- sqrt( (plotdata_scaled$mean_phytoene_scaled - fitted_phytoscaled) ^2)

BCscaledmod <- lm(mean_BC_scaled ~ timept_numeric*layer + genotypeID, data=plotdata_scaled)
fitted_BCscaled <- predict(BCscaledmod)
distance_BC <- sqrt( (plotdata_scaled$mean_BC_scaled - fitted_BCscaled) ^2)

DMscaledmod <- lm(mean_DM_scaled ~ timept_numeric*layer + genotypeID, data=plotdata_scaled)
fitted_DMscaled <- predict(DMscaledmod)
distance_DM <- sqrt( (plotdata_scaled$mean_DM_scaled - fitted_DMscaled) ^2)

genoranks <- cbind(plotdata_scaled[,c("genotypeID", "timept_numeric", "layer")], 
                   distance_DM = distance_DM,
                   distance_TC = distance_TC,
                   distance_phytoene = distance_phytoene,
                   distance_BC = distance_BC) %>%
  group_by(genotypeID) %>%
  summarise(distance_DM = mean(distance_DM)/sd(distance_DM), 
            distance_TC = mean(distance_TC)/sd(distance_TC), 
            distance_phytoene = mean(distance_phytoene)/sd(distance_phytoene), 
            distance_BC = mean(distance_BC)/sd(distance_BC),
            total_phyto_DM_distance = sum(distance_DM, distance_phytoene)) %>%
  filter(genotypeID %in% complete_obs_genos$genotypeID) %>% #select only genotypes that have all samples present
  arrange(distance_DM) %>% mutate(DM_rank = 1:nrow(.)) %>%
  arrange(distance_TC) %>% mutate(TC_rank = 1:nrow(.)) %>%
  rowwise() %>% mutate(avg_rank = mean(c(DM_rank, TC_rank))) %>% ungroup() %>%
  arrange(avg_rank) %>% 
  slice_head(n=4)


plotdata <- plotdata %>% mutate(selected = 
                                  case_when(genotypeID %in% genoranks$genotypeID ~ "yes", T ~ "no"))

ggplot(plotdata, aes(timept_numeric, mean_TC)) +
  geom_point(aes(shape=layer)) +
  geom_smooth(data = plotdata %>% filter(layer=="inner"), aes(lty = "inner"), method = "lm", linewidth = 0.5, color="darkgray", se=F) +
  geom_smooth(data = plotdata %>% filter(layer=="outer"), aes(lty = "outer"), method = "lm", linewidth = 0.5, color="darkgray", se=F) +
  scale_linetype_manual(name = "Linear model layer", values = c("inner" = "dashed", "outer" = "solid")) +
  geom_point(data = plotdata %>% filter(selected=="yes"), aes(color=genotypeID, shape=layer), size=3) +
  xlab("time point (MAP)") + ylab("mean TC") +
  
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"] + coef(TCscaledmod)["layerouter"] + coef(TCscaledmod)["genotypeIDX1_54"], 
              slope = coef(TCscaledmod)["timept_numeric"] + coef(TCscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#7CAE00", lty = "solid") +
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"] + coef(TCscaledmod)["genotypeIDX1_54"],
              slope = coef(TCscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#7CAE00", lty = "dashed") +
  
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"] + coef(TCscaledmod)["layerouter"] + coef(TCscaledmod)["genotypeIDC17"],
              slope = coef(TCscaledmod)["timept_numeric"] + coef(TCscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#F8766D", lty = "solid") +
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"] + coef(TCscaledmod)["genotypeIDC17"],
              slope = coef(TCscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#F8766D", lty = "dashed") +
  
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"] +  coef(TCscaledmod)["layerouter"] + coef(TCscaledmod)["genotypeIDX11_68"] ,
              slope = coef(TCscaledmod)["timept_numeric"] + coef(TCscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#00BFC4", lty = "solid") +
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"] + coef(TCscaledmod)["genotypeIDX11_68"],
              slope = coef(TCscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#00BFC4", lty = "dashed") +
  
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"]  + coef(TCscaledmod)["layerouter"] + coef(TCscaledmod)["genotypeIDX2_24"],
              slope = coef(TCscaledmod)["timept_numeric"]+ coef(TCscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#C77CFF", lty = "solid") +
  geom_abline(intercept = coef(TCscaledmod)["(Intercept)"] + coef(TCscaledmod)["genotypeIDX2_24"],
              slope = coef(TCscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#C77CFF", lty = "dashed")
  


ggplot(plotdata, aes(timept_numeric, mean_DM)) +   
  geom_point(aes(shape=layer)) +
  geom_smooth(data = plotdata %>% filter(layer=="inner"), aes(lty = "inner"), method = "lm", linewidth = 0.5, color="darkgray", se=F) +
  geom_smooth(data = plotdata %>% filter(layer=="outer"), aes(lty = "outer"), method = "lm", linewidth = 0.5, color="darkgray", se=F) +
  geom_point(data = plotdata %>% filter(selected=="yes"), aes(color=genotypeID, shape=layer), size=3) +
  scale_linetype_manual(name = "Linear model layer", values = c("inner" = "dashed", "outer" = "solid")) +
  xlab("time point (MAP)") + ylab("mean DM") +
    geom_abline(intercept = coef(DMscaledmod)["(Intercept)"] + coef(DMscaledmod)["layerouter"] + coef(DMscaledmod)["genotypeIDX1_54"], 
              slope = coef(DMscaledmod)["timept_numeric"] + coef(DMscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#7CAE00", lty = "solid") +
  geom_abline(intercept = coef(DMscaledmod)["(Intercept)"] + coef(DMscaledmod)["genotypeIDX1_54"],
              slope = coef(DMscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#7CAE00", lty = "dashed") +
  
  geom_abline(intercept = coef(DMscaledmod)["(Intercept)"] + coef(DMscaledmod)["layerouter"] + coef(DMscaledmod)["genotypeIDC17"],
              slope = coef(DMscaledmod)["timept_numeric"] + coef(DMscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#F8766D", lty = "solid") +
  geom_abline(intercept = coef(DMscaledmod)["(Intercept)"] + coef(DMscaledmod)["genotypeIDC17"],
              slope = coef(DMscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#F8766D", lty = "dashed") +
  
  geom_abline(intercept = coef(DMscaledmod)["(Intercept)"] +  coef(DMscaledmod)["layerouter"] + coef(DMscaledmod)["genotypeIDX11_68"] ,
              slope = coef(DMscaledmod)["timept_numeric"] + coef(DMscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#00BFC4", lty = "solid") +
  geom_abline(intercept = coef(DMscaledmod)["(Intercept)"] + coef(DMscaledmod)["genotypeIDX11_68"],
              slope = coef(DMscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#00BFC4", lty = "dashed") +
  
  geom_abline(intercept = coef(DMscaledmod)["(Intercept)"]  + coef(DMscaledmod)["layerouter"] + coef(DMscaledmod)["genotypeIDX2_24"],
              slope = coef(DMscaledmod)["timept_numeric"]+ coef(DMscaledmod)["timept_numeric:layerouter"],
              linewidth = 0.7, color="#C77CFF", lty = "solid") +
  geom_abline(intercept = coef(DMscaledmod)["(Intercept)"] + coef(DMscaledmod)["genotypeIDX2_24"],
              slope = coef(DMscaledmod)["timept_numeric"],
              linewidth = 0.7, color="#C77CFF", lty = "dashed")

ggplot(plotdata, aes(timept_numeric, mean_phytoene)) +
  geom_point(aes(shape=layer)) +
  geom_smooth(data = plotdata %>% filter(layer=="inner"), aes(lty = "inner"), method = "lm", linewidth = 0.7, color="black", se=F) +
  geom_smooth(data = plotdata %>% filter(layer=="outer"), aes(lty = "outer"), method = "lm", linewidth = 0.7, color="black", se=F) +
  geom_point(data = plotdata %>% filter(selected=="yes"), aes(color=genotypeID, shape=layer), size=3) +
  scale_linetype_manual(name = "layer", values = c("inner" = "dashed", "outer" = "solid")) +
  xlab("time point (MAP)") + ylab("mean phytoene")

ggplot(plotdata, aes(timept_numeric, mean_BC)) +
  geom_point(aes(shape=layer)) +
  geom_smooth(data = plotdata %>% filter(layer=="inner"), aes(lty = "inner"), method = "lm", linewidth = 0.7, color="black", se=F) +
  geom_smooth(data = plotdata %>% filter(layer=="outer"), aes(lty = "outer"), method = "lm", linewidth = 0.7, color="black", se=F) +
  geom_point(data = plotdata %>% filter(selected=="yes"), aes(color=genotypeID, shape=layer), size=3) +
  scale_linetype_manual(name = "layer", values = c("inner" = "dashed", "outer" = "solid")) +
  xlab("time point (MAP)") + ylab("mean beta-carotene")

```


  
  
  
  

#T tests for inner/outer layer

```{r}
t.test(inner_df$total_carotenoids, outer_df$total_carotenoids)
# significantly less total carotenoids in outer layers compared to inner

t.test(inner_df$dry_matter, outer_df$dry_matter)
# no significant difference between DM content in inner vs outer layers 

t.test(inner_df$Beta.Carotene, outer_df$Beta.Carotene)
# significantly less beta carotene in outer layers compared to inner

t.test(inner_df$Phytoene, outer_df$Phytoene)
# significantly less phytoene in outer layers compared to inner

```



#Linear Models 

##model_1 : effect of carotenoids on DM 
```{r}
model_1 <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept)
summary(model_1)
plot(joined_df$total_carotenoids, joined_df$dry_matter) #check linear relationship
hist(residuals(model_1))

```
- increasing TC by one ug/g has a significant negative effect on DM content (-0.05 %) after accounting for root layer, block number, and timepoint. 
- after accounting for TC, root layer, and block number, there is a significant decrease in DM content across timepoints.
- layer doesn't significantly contribute to the variation in DM content after accounting for TC, block number, and timepoint 

##model_2 : effect of DM on TC
```{r}
model_2 <- lm(data = joined_df, total_carotenoids ~ dry_matter + layer + block + timept)
summary(model_2)
hist(residuals(model_2))
```
- increasing DM by 1% reduces TC content by -0.68 ug/g. DM has a significant negative effect on TC after accounting for root layer, block number, and timepoint.
- TC content is significantly reduced in the outer layer compared to the inner layer after accounting for DM content, block number, and timepoint. 

##pairwise correlation of carotenoids 
```{r}
cor(joined_df %>% dplyr::select(Phytoene:Lutein, total_carotenoids, total_violaxanthin), use = "pairwise.complete.obs")


# average % beta carotene:
mean(joined_df$Beta.Carotene/joined_df$total_carotenoids, na.rm=T)
2 * sd(joined_df$Beta.Carotene/joined_df$total_carotenoids, na.rm=T)

```
- Violaxanthins have low correlations to other some carotenoids, and are slightly negatively correlated with zeta carotene. 

##model_3 : effect of DM on total_violaxanthin 
```{r}
model_3 <- lm(data = joined_df, total_violaxanthin ~ dry_matter + layer + block + timept + total_carotenoids)
summary(model_3)
#cor.test(joined_df$total_violaxanthin, joined_df$total_carotenoids)

```
- After accounting for total carotenoids & experimental factors, DM has a positive relationship with total violas.


##model_4 : effect on DM, TC:Layer
- is the effect of TC on DM dependent of the root layer?
```{r}
model_4 <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + total_carotenoids:layer + block + timept)
summary(model_4)

#F test with/without interaction term 
model_4test <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept)
anova_TClayer_DM <- anova(model_4test, model_4)
anova_TClayer_DM
```
- the interaction between carotenoids and layer is (marginally) important in predicting dry matter. 

### plot effect of TC on DM across layers 

```{r}
summary(model_4)$coefficients

#B of tc:layer on DM from model_4
tclayerinner <- model_4$coefficients["total_carotenoids"]
tclayerouter <- model_4$coefficients["total_carotenoids:layerouter"]

#adding tclayerinner value, right now taking it as = 0 
tclayers <- c(tclayerinner, tclayerouter+tclayerinner)

#create df with new column inner and outer 
tclayers_df <- data.frame(tclayers, layers = c("inner", "outer"))

#extracting standard errors from model_4
se_tclayerinner <- summary(model_4)$coefficients["total_carotenoids",2]
se_tclayerouter <- summary(model_4)$coefficients["total_carotenoids:layerouter",2]

#variances from model_4
var_tclayerinner <- se_tclayerinner^2
var_tclayerouter <- se_tclayerouter^2

#FIXING VARIANCES 

#getting covariances 
tclayers_covariances <- vcov(model_4)[2,c("total_carotenoids:layerouter")]

#fixing the variances summing the var of tclayerinner
#formula: var(x+y) = var(x)+var(y)+2cov(x,y)
varfix_tclayerinner <- var_tclayerinner
varfix_tclayerouter <- var_tclayerinner + var_tclayerouter +
  2*tclayers_covariances

#standard errors fixed
sefix_tclayerinner <- se_tclayerinner
sefix_tclayerouter <- sqrt(varfix_tclayerouter)

std_errorslayers <- c(sefix_tclayerinner, sefix_tclayerouter)

tclayers_df <- tclayers_df %>% 
  mutate(std_errors = std_errorslayers)

#plot 
ggplot(data = tclayers_df)+
  aes(x = layers, y = tclayers)+
  geom_point()+
  geom_errorbar(
    aes(ymin = tclayers - 2*std_errors,
        ymax = tclayers + 2*std_errors, width=.2,
        ))


```

- the effect of carotenoids on DM  is less negative in the outer layers.


### model_4 effect on DM, TC:layer with genotypes 

```{r}
#effect of TC:layer on DM + genotypes  
model_4B <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + total_carotenoids:layer + block + timept + genotypeID)
summary(model_4B)

#without TC:layer but with genotype 
model_4Btest <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept + genotypeID)

#F test 
anova_TClayergen <- anova(model_4B, model_4Btest)
anova_TClayergen

#now with genotype as random effect 
model_4C <- lmer(data = joined_df, dry_matter ~ total_carotenoids + layer + total_carotenoids:layer + block + timept + (1|genotypeID))

model_4Ctest <- lmer(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept + (1|genotypeID))

anova_4C<- anova(model_4C, model_4Ctest)
anova_4C

```




##model_5 : effect on  DM, TC:Timept
- is the effect of TC on DM dependent on timepoint?

```{r}
model_5 <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept + total_carotenoids:timept)
summary(model_5)

#F test with/wihtout interaction term 
model_5test <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept)
summary(model_5test)

anova_5 <- anova(model_5, model_5test)
anova_5


#just TC
model_i <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept)
summary(model_i)

```
The interaction between TC and time point is not significant in the F-test


###plot effect of TC on DM over timepoint
```{r}
summary(model_5)$coefficients

#B of tc:timpt on DM from model_5
tctimept1 <- model_5$coefficients["total_carotenoids"]
tctimept2 <- model_5$coefficients["total_carotenoids:timept7_MAP"]
tctimept3 <- model_5$coefficients["total_carotenoids:timept8_MAP"]  
tctimept4 <- model_5$coefficients["total_carotenoids:timept9_MAP"]

tctimepts <- c(tctimept2, tctimept3, tctimept4)
#add TC value, currently is taking tc:timept1 as = 0
tc_tctimepts <- c(tctimept1, tctimepts + tctimept1)

#create df with new column 1:4 
tc_tctimepts_df <- data.frame(tc_tctimepts, timept = c(1:4))

#extracting standard errors from model_5
se_tctimept1 <- summary(model_5)$coefficients["total_carotenoids",2]
se_tctimept2 <- summary(model_5)$coefficients["total_carotenoids:timept7_MAP",2]
se_tctimept3 <- summary(model_5)$coefficients["total_carotenoids:timept8_MAP",2]
se_tctimept4 <- summary(model_5)$coefficients["total_carotenoids:timept9_MAP",2]

#variances from model_5
var_tctimept1 <- se_tctimept1^2
var_tctimept2 <- se_tctimept2^2
var_tctimept3 <- se_tctimept3^2
var_tctimept4 <- se_tctimept4^2

#FIXING VARIANCES 

#getting covariances 
tccovariances <- vcov(model_5)[2,c("total_carotenoids:timept7_MAP","total_carotenoids:timept8_MAP", "total_carotenoids:timept9_MAP")]

#fixing the variances summing the var of tctimept1
#formula: var(x+y) = var(x)+var(y)+2cov(x,y)
varfix_tctimept1 <- var_tctimept1
varfix_tctimept2 <- var_tctimept1 + var_tctimept2 +
  2*tccovariances["total_carotenoids:timept2"]
varfix_tctimept3 <- var_tctimept1 + var_tctimept3 +
  2*tccovariances["total_carotenoids:timept3"]
varfix_tctimept4 <- var_tctimept1 + var_tctimept4 +
  2*tccovariances["total_carotenoids:timept4"]

#standard errors fixed
sefix_tctimept1 <- se_tctimept1
sefix_tctimept2 <- sqrt(varfix_tctimept2)
sefix_tctimept3 <- sqrt(varfix_tctimept3)
sefix_tctimept4 <- sqrt(varfix_tctimept4)

std_errors <- c(sefix_tctimept1, sefix_tctimept2, sefix_tctimept3,
                sefix_tctimept4)

tc_tctimepts_df <- tc_tctimepts_df %>%
  mutate(std_errors = std_errors)


#plot 
ggplot(data = tc_tctimepts_df)+
  aes(x = timept, y = tc_tctimepts)+
  geom_point()+
  geom_errorbar(
    aes(ymin = tc_tctimepts - 2*std_errors,
        ymax = tc_tctimepts + 2*std_errors, width=.2,
        ))+
  labs(x = "Timepoint",
       y = "Estimate coefficients")

```

- the effect of TC on DM becomes more negative in later timepoints of development after accounting for the effect of root layer and block.




##model_6 :  effect of DM on TC, with DM:Layer
- is the effect of DM on TC dependent of layer? 
```{r}
model_6 <- lm(data = joined_df, total_carotenoids ~ dry_matter*layer +  block + timept)
summary(model_6)
```

##model_7 : effect of DM on TC, DM:Timept
- is the effect of DM on TC dependent of timepoint? 
```{r}
model_7 <- lm(data = joined_df,  total_carotenoids ~ dry_matter + layer + block + timept + dry_matter*timept)
summary(model_7)
```





#Mixed Model: 
##Genotype categories as random effect in DM 

```{r}
mixmod_1 <- lmer(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept + (1 | genotype.category))
summary(mixmod_1)

```
###Getting the pvalue of mixed model (genotype categories) 

```{r}
mixmod_1pval <- lmerTest::lmer(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept + (1 | genotype.category))

summary(mixmod_1pval)

```

##Genotypes as random effect on DM 

```{r}
#make mixed model with genotypes (not category and no total carotenoids)
mixmod_2 <- lmer(data = joined_df, dry_matter ~  layer + block + timept + (1 | genotypeID))
summary(mixmod_2)
```


###DM BLUPS
- extract random effects from the mixed model (genotype as random effect on DM)

```{r}
DM_BLUPS <- ranef(mixmod_2) 

DM_BLUPS
```

#F-Tests 
- to determine if certain terms are necessary in our models. We create full and reduced models and compare them with anova tests.

##model_8 :  effect of technician on TC
```{r}
#this will be the "full" model to model_2
model_8 <- lm(data = joined_df, total_carotenoids ~ dry_matter + layer + block + timept + technician)
summary(model_8)

#F-test 
anova_technicianTC <- anova(model_8, model_2)
anova_technicianTC
```
- technician does not significantly improve the model fit for TC

##model_9 : effect of block on TC
```{r}
#this will be the "reduced" model to model_2 
model_9 <- lm(data = joined_df, total_carotenoids ~ dry_matter + layer + timept)
summary(model_9)

#F-test
anova_blockTC <- anova(model_9, model_2)
anova_blockTC
```
- block does not significantly improve the model fit for TC 

##model_10 : effect of block on DM
```{r}
#this will be the "reduced" model to model_1 
model_10 <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + timept)
summary(model_10)

#F-test 
anova_blockDM <- anova(model_10, model_1)
anova_blockDM
```

- block does significantly improve the model fit for DM 


## model_11 : effect of timepoint in TC 
```{r}
#this will be the "reduced" model to model_2
model_11 <- lm(data = joined_df, total_carotenoids ~ dry_matter + layer + block)

#F-test
anova_timeptTC <- anova(model_11, model_2)
anova_timeptTC

```
- timepoint does significantly improve the model fit for TC 


## model_12 : effect of timepoint in DM 
```{r}
#this will be the "reduced" model to model_1
model_12 <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block)

#F-test
anova_timeptTC <- anova(model_12, model_1)
anova_timeptTC
```
- timepoint does significantly improve the model fit for DM



## model_13 : effect of genotypes in DM 

```{r}
df_genotypes <- joined_df %>%
  dplyr::select(dry_matter, total_carotenoids, layer, block, timept, genotypeID) %>% na.omit()

model_13 <- lm(data = df_genotypes, dry_matter ~ total_carotenoids + layer +
            block + timept)
model_14 <- lm(data = df_genotypes, dry_matter ~ total_carotenoids + layer +
            block + timept + genotypeID)

summary(model_13)
summary(model_14)

#F-test
anova_genotypeDM <- anova(model_13, model_14)
anova_genotypeDM
```

## model_15: effect of phytoene in DM
- to help test the hypothesis that phyotene accumulating may be promoting the degradation or inhibition of starch by the change of amyloplasts to chromoplasts

```{r}
#this is the "full" model to model_1 
model_15 <- lm(data = joined_df, dry_matter ~ total_carotenoids + layer + block + timept + Phytoene)

summary(model_15)


#F test 
anova_phytoeneDM <- anova(model_1, model_15)
anova_phytoeneDM

#cor.test
cor.test(joined_df$Phytoene, joined_df$dry_matter)

t.test(inner_df$Phytoene, outer_df$Phytoene)

```
- phytoene does improve the model's fit to explain the variation of DM even after accounting for the effect of TC on DM. 

## model_16: genotype as random effect but with/wo TC 

```{r}
model_16 <- lmer(data = df_genotypes, dry_matter ~ total_carotenoids + layer + block + timept + (1|genotypeID))
summary(model_16)

model_16test <- lmer(data = df_genotypes, dry_matter ~ layer + block + timept + (1|genotypeID))

anova_16 <- anova(model_16, model_16test)
anova_16
```
TC does not have a significantly negative effect on DM after accounting for genotype.
(Evidence against (?) the hypothesis of a physiological, non-genetic trade off)

#Principal Component Analysis (PCA)
```{r}

#removing NAs for PCA 
clean_joined_df<- joined_df %>% drop_na(3:9) %>% filter(!is.na(genotype.category))
colnames(clean_joined_df)[3:9]

#performing PCA on carotenoids 
pca <- prcomp(clean_joined_df[,3:9], scale = TRUE)
pca
summary(pca)

#percentage of variation each PC accounts for in the original data 
pca.var <- pca$sdev^2 
pca.var.per <- round(pca.var / sum(pca.var)*100, 1) 


#Scree Plot 
barplot(pca.var.per, 
        main = "Scree Plot", 
        xlab = "Principal Component", 
        ylab = "Percent Variation")


```
- PC1 and PC2 combined account for ~87% of the variance in the original data.

##PCA plot clustering by layer 
```{r}
library(ggfortify)

#building data frame 
pca.data <- as.data.frame(pca$x[, 1:2]) # extract first two PCs
pca.data.layer <- cbind(pca.data, clean_joined_df$layer) # add layer to df
colnames(pca.data.layer) <- c("PC1", "PC2", "layer") #change column names 

#plot 
ggplot(pca.data.layer)+
  aes(PC1, PC2, color = layer, shape = layer)+
  geom_point()+
  coord_fixed()+ 
  labs(x = paste0("PC1: ", summary(pca)$importance[2,1]*100, "%"),
       y = paste0("PC2: ", summary(pca)$importance[2,2]*100, "%"))

#plot with ggfortify
pca.loadings <- pca$rotation
autoplot(pca, data = clean_joined_df, color = 'layer', loadings = TRUE, loadings.label = 1)


       
```
- zeta, phytoene, phytophluene, and beta have the most influence in PC1, account for majority of the variance in the original data ??


##PCA plot clustering by genotype
```{r}
#building data frame 
pca.data.genotype <- cbind(pca.data, clean_joined_df$genotypeID) # add genotype to df
colnames(pca.data.genotype) <- c("PC1", "PC2", "genotypeID") #change column names 

#plot 
ggplot(pca.data.genotype)+
  aes(PC1, PC2, color = genotypeID)+
  geom_point()+
  coord_fixed()+ 
  labs(x = paste0("PC1: ", summary(pca)$importance[2,1]*100, "%"),
       y = paste0("PC2: ", summary(pca)$importance[2,2]*100, "%"))


# Get loadings and manually place labels
loadings_df <- as.data.frame(pca$rotation[, 1:2])
loadings_df$varname <- rownames(loadings_df)
loadings_df$PC2[loadings_df$varname=="Violaxanthin.peak.2"] <- loadings_df$PC2[loadings_df$varname=="Violaxanthin.peak.2"]*1.1 #nudge up slightly to prevent overlapping labels



#plot with ggfortify
autoplot(pca, data = clean_joined_df, color = 'genotype.category', 
         loadings = TRUE, loadings.label = F, 
         loadings.col = "black", loadings.label.col ="black", loadings.legend.name="test") +
  geom_text(data = loadings_df, aes(x = PC1*0.3, y = PC2*0.3, label = varname),
            size = 3, hjust=1.1, color = "black") +
  scale_color_discrete(name = "Genotype category")

ggsave("plot_pca_genotype_fixedlabels.png", width = 8.5, height = 6.5, dpi = 400)

```

###PCA plot clustering by genotype (trying to group African, Latin American, & Hybrids)
```{r}
clean_joined_df <- clean_joined_df %>% 
  mutate(genotype.category = case_when(
    startsWith(genotypeID, "A") ~ "African",
    startsWith(genotypeID, "C") ~ "Latin American",
    startsWith(genotypeID, "X") ~ "Hybrid"))

pca.data.genotypecat <- cbind(pca.data, clean_joined_df$genotype.category) # add genotype categories to df
colnames(pca.data.genotypecat) <- c("PC1", "PC2", "genotype.category") #change column names 

#plot 
ggplot(pca.data.genotypecat)+
  aes(PC1, PC2, color = genotype.category)+
  geom_point()+
  coord_fixed()+ 
  labs(x = paste0("PC1: ", summary(pca)$importance[2,1]*100, "%"),
       y = paste0("PC2: ", summary(pca)$importance[2,2]*100, "%"))
 stat_ellipse(geom="polygon", level=0.95, alpha=0.2) #adding a stat
 
 

#plot with ggfortify
plot_pca_genotype <- autoplot(pca, data = clean_joined_df, color = 'genotype.category', loadings = TRUE, loadings.label = F)+
  labs(color = "Genotype")

plot_pca_genotype
ggsave("plot_pca_genotype.png", width = 8.5, height = 6.5, dpi = 400)

```
- to read: ej. zeta carotene is negatively correlated to lutein? Lutein and viola closely related.
- ?? zeta, beta, phytoene and phytofluene influence PC1 while lutein and violas PC2?

### model_17: PC1 and PC2 on DM 

```{r}
clean_joined_df <- clean_joined_df %>% 
  mutate(PC1 = pca$x[,1],
         PC2 = pca$x[,2])

model_17 <- lmer(data = clean_joined_df, dry_matter ~ layer + block + timept + PC1 + PC2 + (1|genotypeID)) 
summary(model_17)


```

## model_18: effect of timepoint:TC interaction with genotype

```{r}
model_18 <- lmer(data = clean_joined_df, dry_matter ~ timept*total_carotenoids + layer + block + (1|genotypeID))
summary(model_18)

model_18test <- lmer(data = clean_joined_df, dry_matter ~ total_carotenoids + timept + layer + block + (1|genotypeID))

anova_18 <- anova(model_18, model_18test)
anova_18
```


## model_19: effect of cis-carotenes vs. total carotenoids

```{r}
clean_joined_df <- clean_joined_df %>%
  mutate(totalciscarotenes = Phytoene + Phytofluene + Zeta.Carotene, 
         totalxanthophylls = total_violaxanthin + Lutein,
         totalothercarotenoids = total_violaxanthin + Lutein + Beta.Carotene)


model_19 <- lmer(data = clean_joined_df, dry_matter ~ totalciscarotenes + totalothercarotenoids + timept + layer + block + (1|genotypeID), REML=FALSE)
summary(model_19)

model_19test <- lmer(data = clean_joined_df, dry_matter ~ totalothercarotenoids + timept + layer + block + (1|genotypeID), REML=FALSE)

anova_19 <- anova(model_19, model_19test)
anova_19 #p=0.000339
```


## model_19.2: effect of lutein vs. total_carotenoids 

```{r}
clean_joined_df <- clean_joined_df %>%
  mutate(totalciscarotenes = Phytoene + Phytofluene + Zeta.Carotene, 
         totalxanthophylls = total_violaxanthin + Lutein,
         totalothercarotenoids = total_violaxanthin + Lutein + Beta.Carotene,
         totalviola = Violaxanthin.peak.1 + Violaxanthin.peak.2,
         nonlutein = Phytoene + Phytofluene + Zeta.Carotene + Beta.Carotene + Violaxanthin.peak.1 + Violaxanthin.peak.2,
         nonphytoene = Phytofluene + Zeta.Carotene + Lutein + Beta.Carotene + Violaxanthin.peak.1 + Violaxanthin.peak.2,
         nonphytofluene = Phytoene + Zeta.Carotene + Lutein + Beta.Carotene + Violaxanthin.peak.1 + Violaxanthin.peak.2,
         nonzeta = Phytoene + Phytofluene + Lutein + Beta.Carotene + Violaxanthin.peak.1 + Violaxanthin.peak.2,
         nonbeta = Phytoene + Phytofluene + Zeta.Carotene + Lutein + Violaxanthin.peak.1 + Violaxanthin.peak.2,
         nonviola = Phytoene + Phytofluene + Zeta.Carotene + Lutein + Beta.Carotene)


# Lutein vs all other carotenoids
model_19_2 <- lmer(data = clean_joined_df, dry_matter ~ Lutein + nonlutein + timept + layer + block + (1|genotypeID), REML=FALSE)
summary(model_19_2)

model_19_2_test <- lmer(data = clean_joined_df, dry_matter ~ nonlutein + timept + layer + block + (1|genotypeID), REML=FALSE)
anova_19_2 <- anova(model_19_2, model_19_2_test)
anova_19_2 # X2 = 0.1666, p = 0.6832

# Phytoene vs all other carotenoids
model_19_2 <- lmer(data = clean_joined_df, dry_matter ~ Phytoene + nonphytoene + timept + layer + block + (1|genotypeID), REML=FALSE)
summary(model_19_2)

model_19_2_test <- lmer(data = clean_joined_df, dry_matter ~ nonphytoene + timept + layer + block + (1|genotypeID), REML=FALSE)
anova_19_2 <- anova(model_19_2, model_19_2_test)
anova_19_2 # X2 =  17.028, p = 5.662e-05


# Phytofluene vs all other carotenoids
model_19_2 <- lmer(data = clean_joined_df, dry_matter ~ Phytofluene + nonphytofluene + timept + layer + block + (1|genotypeID), REML=FALSE)
summary(model_19_2)

model_19_2_test <- lmer(data = clean_joined_df, dry_matter ~ nonphytofluene + timept + layer + block + (1|genotypeID), REML=FALSE)
anova_19_2 <- anova(model_19_2, model_19_2_test)
anova_19_2 # X2= 10.211, p=0.001396


# Zeta carotene vs all other carotenoids
model_19_2 <- lmer(data = clean_joined_df, dry_matter ~ Zeta.Carotene + nonzeta + timept + layer + block + (1|genotypeID), REML=FALSE)
summary(model_19_2)

model_19_2_test <- lmer(data = clean_joined_df, dry_matter ~ nonzeta + timept + layer + block + (1|genotypeID), REML=FALSE)
anova_19_2 <- anova(model_19_2, model_19_2_test)
anova_19_2 # X2=1.269, p = 0.2598


# Beta carotene vs all other carotenoids
model_19_2 <- lmer(data = clean_joined_df, dry_matter ~ Beta.Carotene + nonbeta + timept + layer + block + (1|genotypeID), REML=FALSE)
summary(model_19_2)

model_19_2_test <- lmer(data = clean_joined_df, dry_matter ~ nonbeta + timept + layer + block + (1|genotypeID), REML=FALSE)
anova_19_2 <- anova(model_19_2, model_19_2_test)
anova_19_2 #X2=2.0955, p = 0.1477



# Violaxanthin vs all other carotenoids
model_19_2 <- lmer(data = clean_joined_df, dry_matter ~ totalviola + nonviola + timept + layer + block + (1|genotypeID), REML=FALSE)
summary(model_19_2)

model_19_2_test <- lmer(data = clean_joined_df, dry_matter ~ nonviola + timept + layer + block + (1|genotypeID), REML=FALSE)
anova_19_2 <- anova(model_19_2, model_19_2_test)
anova_19_2 #X2=1.33, p = 0.2488
```




#Phytoene:TC relationship with timepoint

```{r}
clean_dftimept <- clean_joined_df %>% group_by(timept) %>%
  summarise(phy_ratio = mean(Phytoene/total_carotenoids),
            phy_ratio_sd = sd(Phytoene/total_carotenoids))


ggplot(data = clean_dftimept)+
  aes(x = timept, y = phy_ratio)+
  geom_col()+
  geom_errorbar(aes(ymin = phy_ratio - 2*phy_ratio_sd,
    ymax = phy_ratio + 2*phy_ratio_sd))


clean_dftimept2 <- clean_joined_df %>% group_by(timept) %>%
  summarise(PPZB_ratio = mean(Phytoene+Phytofluene+Zeta.Carotene+Beta.Carotene/total_violaxanthin+Lutein),
            PPZB_ratio_sd = sd(Phytoene+Phytofluene+Zeta.Carotene+Beta.Carotene/total_violaxanthin+Lutein))
            

ggplot(data = clean_dftimept2)+
  aes(x = timept, y = PPZB_ratio)+
  geom_col()+
  geom_errorbar(aes(ymin = PPZB_ratio - 2*PPZB_ratio_sd,
    ymax = PPZB_ratio + 2*PPZB_ratio_sd))


```



#TC/DM relationship by tissue layer 

```{r}
plot_TCDM_layer <- ggplot(data = joined_df %>% filter(!is.na(layer)))+
  aes(x = total_carotenoids, y = dry_matter, color = layer)+
  geom_point(alpha=0.8)+
  geom_smooth(method = lm, linetype = "dashed", se = FALSE, fullrange = TRUE)+
  xlab("Total carotenoids (μg/g DW)")+
  ylab("Dry matter (%)") + 
  ggtitle("Year 2 trial")


plot_TCDM_layer
ggsave("plot_TCDM_layer.png", width = 8.5, height = 6.5, dpi = 400)

```

#TC/DM relationship by timepoints

```{r}
plot_TCDM_timept <- ggplot(data = joined_df %>% filter(dry_matter < 55))+
  aes(x = total_carotenoids, y = dry_matter, color = timept)+
  geom_point()+
  geom_smooth(method = lm, linetype = "dashed", se = FALSE, fullrange = TRUE)+
  xlab("Total carotenoids (μg/g DW)")+
  ylab("Dry matter (%)")+
  labs(color = "timepoint")+
  scale_color_discrete(labels = c('6 MAP', '7 MAP', '8 MAP', '9 MAP', 'NA'))

plot_TCDM_timept
ggsave("plot_TCDM_timept.png", width = 8.5, height = 6.5, dpi = 400)
```

#TC/DM relationship by genotype categories 
```{r}

plot_TCDM_genotypecat <- ggplot(data = joined_df)+
  aes(x = total_carotenoids, y = dry_matter, color = genotype.category)+
  geom_point()+
  geom_smooth(method = lm, linetype = "dashed", se = FALSE, fullrange = TRUE)+
  xlab("Total carotenoids (μg/g)")+
  ylab("Dry matter (%)")+
  labs(color = "Genotype")

plot_TCDM_genotypecat
ggsave("plot_TCDM_genotypecat.png", width = 8.5, height = 6.5, dpi = 400)

```
#Boxplots
##TC over timepoint

```{r}
ggplot(data = joined_df)+
  aes(x = timept, y = total_carotenoids)+
  geom_boxplot()
```
##DM over timepoints
```{r}
ggplot(joined_df)+
  aes(x = timept, y = dry_matter)+
  geom_boxplot()
```
## DM based on genotype categories??

```{r}
ggplot(joined_df)+
  aes(x = genotype.category, y = dry_matter)+
  geom_point()

```
#Carotenoids over timepoint (all of them)

```{r}
multivariate_model1 <- lm(cbind(Violaxanthin.peak.1, Violaxanthin.peak.2, Lutein) ~ timept, data = joined_df)
summary(multivariate_model1)

multivariate_model2 <- lm(cbind(Phytoene, Phytofluene, Zeta.Carotene, Beta.Carotene) ~ timept, data = joined_df)
summary(multivariate_model2)
#This model will assess the effect of timept on all three response variables simultaneously. 
```

# Heritability 
- broad sense heritability (we're using genotype as a factor, includes total genetic variance)
- formula (Vg/Vg + Ve)

```{r}
her_modelDM <- lmer(data = joined_df, dry_matter ~ layer + block + timept + (1|genotype))

her_modelTC <- lmer(data = joined_df, total_carotenoids ~ layer + block + timept + (1|genotype))

summary(her_modelDM)
heritabilityDM <- as.data.frame((summary(her_modelDM))$varcor)^2
summary(her_modelTC) 
```


Heritability of the egg-yolk phenotype:
```{r}
phenodata_merged_eggyolk = readRDS("../RNAseq_yr2/data/Carotenoids_DM_phenodata_yr1-2merged.RDS")


her_modelEY <- lmer(data = joined_df, total_carotenoids ~ layer + block + timept + (1|genotype))
```


Comparison of Latin American vs. African:
```{r}
phenodata_merged <- phenodata_merged_eggyolk %>% mutate(genotype.category = case_when(
    startsWith(genotype, "A") ~ "African",
    startsWith(genotype, "C") ~ "Latin American",
    startsWith(genotype, "X") ~ "Hybrid"))

model_20 <- lmer(data = phenodata_merged, dry_matter ~ total_carotenoids*genotype.category + layer + block + timept + (1|genotype))
summary(model_20)

#total_carotenoids:genotype.categoryLatin American   Estimate: 0.07596 Pr(>|t|):0.19033  

model_20_reduced <- lmer(data = phenodata_merged, dry_matter ~ total_carotenoids + genotype.category + layer + block + timept + (1|genotype))

anova(model_20, model_20_reduced)
# Chisq 9.2448, Df 2, Pr(>Chisq) 0.009829 



### C

```
There is a significant interaction effect between genotype category and total carotenoids on DM (X2 = 9.2448, p=0.009829).



